### 2024-10-11 paper.tex
Initial import

\documentclass[draft,english]{jflart}
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{lstcoq}
\usepackage[numbers]{natbib}
\usepackage{enumitem}

% Numéro et année des JFLAs visées par l'article, obligatoire.
\jfla{36}{2025}

%\title{Functorial Symmetric Cryptography}
\title{Can Symmetric Cryptography Be Liberated from the Von Neumann
  Style?}
\titlerunning{Free Von Neumann!}

% Auteurs
\author[1]{Jean-Baptiste Menant}
\author[2]{Yves Ndiaye}
\author[2,3]{\\Pierre-Évariste Dagand}
\authorrunning{Menant, Ndiaye et Dagand}

% Affiliations des auteurs
\affil[1]{École Normale Supérieure de Lyon}
\affil[2]{Université Paris Cité, IRIF}
\affil[3]{CNRS}

\newcommand{\TODO}[1]{\textcolor{red}{$\langle$ TODO:}#1\textcolor{red}{$\rangle$}}
\newcommand{\ie}{\textit{i.e.}}
\newcommand{\etc}{\textit{etc.}}

\begin{document}

\maketitle

\begin{abstract}

  In defiance of Hinchliffe's rule, this article sets out to
  demonstrate that its title can be answered by the word ``yes''. We
  show that modern symmetric ciphers can be modeled through a small
  set of algebraic structures (Boolean algebra, Naperian and foldable
  functors). We reveal that these ciphers exhibit some interesting
  compositional structure at the type-level, which enables systematic
  code transformation, such as bitslicing and fixslicing. Our work
  rests on a Coq development providing the specification of two
  ciphers, Skinny and Gift, deriving a bitsliced and fixsliced
  implementation for each and proving their correctness with respect
  to their specifications.

\end{abstract}

\section{Introduction}

%% * Amazing cryptographers (ZRR)
%% "trade secret"
%% "genius out of the bottle", like toothpaste
%% ** Section 2: F_{}
%% ** Section 5: Cortex ARM M4 assembly
%% -> Our offer: Section 4
%%   -> shorter Section 5

Anyone who has ever read a cryptography paper knows this for a fact:
cryptographers are at the top of the academic thought chain. In the
archetypal symmetric cryptography paper, there is a ``Section 2''
scene in which cryptographers are bravely waving their most potent
finite field. In a latter ``Section 4'' scene, the very same
cryptographers are seen extolling the virtues of some nifty
\textsc{Arm} Cortex assembly instruction, which takes center stage in
their hand-crafted, manually optimized implementation. No wonder we
jealously stash them away in ``zone à régime restrictive''! The rumor
goes that, behind these padlocked doors, they have developed a set of
domain-specific techniques to deliver very high-throughput software
implementation of symmetric ciphers. This has been slowly gnawing at
compiler writers, for whom job security~\citep{bernstein} has become a
major concern. Luckily for us, every now and then, someone
inadvertently let the genies out of their Klein bottle and we get to
read about the subtle art of ``bitslicing''~\citep{kwan:bitslicing}

The present work aims to cast these two notions in the language of
functional programming. We shall see that bitslicing corresponds to a
run-off-the-mill \emph{data} representation change (in effect, a
matrix transposition). But this is only half of the work: our ability
to systematically transform the \emph{code} operating on such
representation relies on suitably polymorphic definitions:
parametricity at its best! Fixslicing, on the other hand, will be
rationalized through equational reasoning over purely functional
terms. Identifying commuting terms will be of key importance there:
programming with algebraic structures (rather than untamed arrays of
bits) will be instrumental!

In the long run, our ambition is to offer cryptographers the
conceptual apparatus to better illuminate their ``Section 5'' scene.
Each and every symmetric cryptosystem should not have to exhibit a
bitsliced or fixsliced implementation: by just \emph{specifying} the
cryptosystem in the right framework (abstract yet suitably
operational), one should get the guarantee that the design can be
bitsliced and fixsliced. In fact, one would boldly claim that a
compiler could automatically apply these transformations and produce
optimized code, starting from an high-level description of the cipher.
This is already the case for bitslicing~\citep{usuba}, adding support
for fixslicing is next on the list.

%% * Executive summary of modern symmetric cryptography
%% ** iterate pure `round` function N times
%% *** thread a `state`
%% **** initially: clear text
%% **** finally: cipher text
%% *** mixing in constants derived from a secret key
%% **** computed off-line
%% **** off the critical path

As intimidating as cryptographers are, the essence of modern symmetric
cryptography primitives can in fact be distilled in a few paragraph. A
primitive can be understood as a purely functional programs. It takes
as input two blocks of data, one representing the input cleartext and
the other representing the input keys (derived from a single key
through a non-performance critical process called ``key schedule'',
which we shall ignore here). The block size is of order of a hundred
bits (typically, 64, 128 or 256 bits). The output is a block of data
of the same size, the ciphertext.

Looking at the process of turning the cleartext into a ciphertext, one
realizes that cryptographers have been programming in the State monad
all along: ciphers are described as operating on a ``state'',
initialized to be the cleartext and read off as the ciphertext at the
end. This state is updated by a ``round'' function that is repeated a
given number of times (of the order of ten iterations). The exact
number of iterations is a trade-off between latency (less iterations
means faster computation) and security (more iterations makes for more
costly cryptanalysis effort). The round function itself is built
compositionally from 3 components: a key mixing layer (making the
encryption process reversible), a confusion layer (smoothing local
correlations in the inputs) and a diffusion layer (spreading
correlations across the whole output block).

%% ** side-channel attack
%% *** no control-flow
%% **** only statically bounded loops
%% *** no memory access based on data flow
%% **** only register spilling
%% *** -> combinational circuits

To avoid side-channel attack (for example, based on
timing~\citep{time-side-channel}), this pipeline can be more precisely
understood as a combinational circuit. We are forbidden from
performing control-flow operations on secret data, leaving only
statically-bounded loops. We are also forbidden from performing memory
access based on the data flow, leaving only register spilling.

%% ** modernity: starts with AES
%% *** more structured than DES
%% *** in particular: no-trick-up-my-sleeves Sbox
%% ** operate over bit-level n-dimensional array

Following the NSA involvement in the design of the DES diffusion
layer~\citep{des-history}, cryptographers have been advocating for
``nothing-up-my-sleeve'' designs (which does not make magic tricks
impossible, just harder~\citep{bernstein:something-up-my-sleeves}).
For example, the diffusion layer of the AES standard is based on the
usual multiplicative inverse over the finite field induced by an
innocent-looking polynomial. To support such design, the state of
modern ciphers is thus equipped with some matricial structure
(oftentimes, manipulated as a 2D or 3D array of bits): the cipher is
not specified over a flat sequence of individual bits but over a
n-dimensional array of bits.

%% ** designed for high-throughput execution in software 
%% *** -> parallel execution
%% *** -> "scale out" on SIMD

Moreover, ciphers are now, by and large, designed to be run
efficiently on software. The era of hardware-based cryptography is
long gone, even for embedded platforms~\citep{lightweight-crypto}. To
achieve high-throughput, ciphers are typically designed so as to be
admit a bitsliced implementation. Bitslicing enables a form of
parallel execution \emph{within} machine words. We thus gain access to
a form of CPU-level ``scale out'' to larger machine words, in
particular through single-instruction multiple-data (SIMD)
instructions. Doubling the size of machine words yields nearly twice
the throughput, without any effort. Evidences tend to suggest that
cryptographers obtain bitsliced designs through sheer intellectual
might. It is in fact the sole purpose of the ``Section 5'' scene to
exhibit a bitsliced witness in all the gory details.

%% * Contributions
%% ** identify algebraic structure
%% *** mathematical language of software circuits
%% ** rationalize bitslicing
%% *** representation change
%% *** -> key ingredient: parametricity
%% ** rationalize fixslicing
%% *** split & commute permutations
%% *** -> key ingredient: equational reasoning

What would it take to turn this art form into an engineering
principle? We answer through 3 contributions:
%
\begin{itemize}
\item we identify the algebraic structure necessary to model the state
  of a modern symmetric cryptography primitive
  (Section~\ref{sec:spec}). Doing so, we delineate a mathematical
  language of software circuits, rooted in the categorical notion of
  functor to account for data containers~;
\item we give a formal account of bitslicing as a data representation
  change supported by suitably polymorphic definitions
  (Section~\ref{sec:bitslicing}). As expected when it comes to
  switching between data representations, parametricity plays a key
  role to justify the equivalence of the resulting programs~;
\item we give a formal account of fixslicing as a whole-program
  transformation (Section~\ref{sec:fixslicing}). We show how a
  fixsliced implementation can be obtained through equational
  reasoning. To support such reasoning, we crucially rely on the
  algebraic structure set forth in the Section~\ref{sec:spec}.
\end{itemize}

%% ** broader take-away: "functor-oriented programming"
%% *** abstracting over concrete data containers
%% *** modeled as functor + structure (Gibbons)
%% **** -> algebraic reasoning / optimization
%% **** -> turn code back to polymorphic = free theorems
%% ** terminology: "functor" means "categorical functor"
%% *** formally defined in due time

We believe that there is also a broader take-away for an audience of
stubbornly French functional programmers. The present work surfs on
the wake of the Squiggol school of
programming~\citep{wg21,gibbons:naperian}. Because our
government-approved programming language does not yet support ad-hoc
polymorphism~\citep{remy:explicit-poly}, the French community may be
missing out on some interesting programming patterns. The present
article can thus be read as a case-study in ``functor-oriented
programming'', where the word ``functor'' is taken in his categorical
meaning (recalled in Section~\ref{sec:spec} and unrelated to the one
mentioned in the ML module system). In effect, we forbid ourselves
from programming over a concrete datastructure: instead, we rely on
(categorical) functors, peppered with some more structure, so as to 1.
unlock algebraic manipulation during compilation and 2. exploit
parametricity for reasoning.

%% * Purely semantics work
%% ** hence: "maybe"
%% ** /> Coq
%% *** -> executability
%% **** -> testing
%% *** -> prototype compiler
%% **** -> `autorewrite`
%% *** some level of confidence
%% *** past experience with Usuba

The present work is but the beginning of our research program. We
shall focus \emph{exclusively} on the semantics aspects, only briefly
touching upon implementation aspects (which remain the raison d'être
of the project!). Our experience designing the Usuba~\citep{usuba}
programmign language and implementing its compiler gives us some
confidence that a syntax can indeed be tailored to suit this
semantics.

We nonetheless could not resist the temptation of developing the
semantics in the Coq theorem prover, as executable programs. After
all, we are interested in combinational circuits: if there has ever
been a time where we do not have to worry about termination, this is
it! So we have implemented cryptographical primitive in pure Gallina.
This is nothing to worry about: after all, more ambitious colleagues
went from there down to deeply-embedded assembly
code~\citep{chlipala}. Besides the ability to prove the correctness
and, even more usefully, test our code, this also enables us to easily
prototype a fixslicing compiler, using Coq's \texttt{autorewrite}
tactics to emulate a rule-based simplification
engine~\citep{haskell:simplier} (part of an imaginary optimizing
compiler).

%% * Internship work
%% <- JB: BSc on Skinny
%% <- Ndiaye: MSc on Gift
%% -> running examples: Skinny

%% * Disclaimer: none of these "programs" were easy to write / theorems easy to assert
%%   <- don't be fooled by the magician's trick
%%   -> all the more reason to write a language!
%%     -> automatically inhabit the semantics
%%   -> all the more reason to write a compiler!
%%     -> automatically apply theorems

This article grew out of the Bachelor's research project of the first
author and the Master thesis of the second author. The former worked
on the specification, bitslicing and fixslicing of the Skinny
cipher~\citep{skinny-spec, adomin:skinny}. The latter worked on the
specification, bitslicing and fixslicing of the Gift
cipher~\citep{gift-spec, admon:gift}. Both ciphers are provided in the
aaccompanying source code~\footnote{\url{TODO}}. For clarity, we shall
focus exclusively on Skinny as our running example here.

Note that none of the programs below were easy to write in the first
place: it took weeks of careful study to weed out the essential
complexity from the accidental mismanagement of array indices. If the
code seem somewhat trivial, one should bear in mind that this
simplicity was hard-won. All the more reason to, another day, design a
programming language, so as to automatically inhabit this semantics,
and write its compiler, so as to automatically optimize code following
our theorems.

\section{Functional specification}
\label{sec:spec}

%% * Moving beyond "Section 2"
%% ** Operational description
%% ** Non-optimized reference implementation

In this section, we intend to give a specification for the cipher
Skinny, with an eye towards implementation. We shall therefore aim for
a rather operational description, to give a sense of the computational
cost of the cipher, without premature concern for implementation
performance just yet.

%% * Modeling state
%% ** 3 dimensions
%% *** -> figure
%% ** expressed as 3 type formers + (same) structure
%% *** -> introduced incrementally (pedagogy)
%% *** -> summarized in Fig. \ref

As often in functional programming, it helps tremendously to first lay
out the types our program will have to deal with. In our case, the
focal point is the \emph{state} of the cipher, which is specified as a
3-dimensional array (Figure~\ref{fig:state}). We model each dimension
in turn through a dedicated type constructor:
%
\begin{itemize}[nosep]
\item \coqe!Rows.T : Type -> Type! is a data container representing 4 rows of data (horizontal dimension),
\item \coqe!Cols.T : Type -> Type! is a data container representing 4 columns of data (vertical dimension), and
\item \coqe!Slice8.T : Type -> Type! is a data container representing
  8 slices of data (depth dimension).
\end{itemize}

The cube drawn in Figure~\ref{fig:state} could just be modeled as
%
\coqe!Rows.T (Cols.T (Slice8.T bool))!,
%
\coqe!Slice8.T (Rows.T (Cols.T bool))!,
%
or any other composition of these 3 type formers. They are all
isomorphic to a sequence of 128 bits. We shall wait until
Section~\ref{sec:assembling} to settle on the most natural
representation for our specification effort. We revisit this choice in
Section~\ref{sec:bitslicing} when it comes to producing a
memory-efficient representation.

\begin{figure}
  %\includegraphics{skinny_state}
  \caption{Skinny state matrix}
  \label{fig:state}
\end{figure}

%% * Structure: functor
%% ** essence of data container
%% ** contains & preserves collection of individuals
%% ** abstract reading: truely parametric collection (free theorem)

For all intent and purposes, these type constructors shall remain
abstract throughout this article: we interact with them solely through
their algebraic interface. The first of which is the categorical
notion of functor:
%
\coqfrom{Algebra}{Functor}

To the disciples of Reynolds and Girard, this signature has come to
mean ``parametric data container'': how would you go about asserting
that the type constructor \coqe!Rows.T! is not doing something
non-trivial with the type it is provided as an argument? In other
word, how to be sure that a \coqe!Rows.T bool! behaves ``similarly''
as a \coqe!Rows.T nat!? One could try and argue that Coq does not
allow pattern matching on types but it is rather unsavory to involve
the design of the whole programming language into this argument.

Instead, we require \coqe!Rows.T! to offer a \coqe!map! operation: it
constructively witnesses the fact that a \coqe!Rows.T A! is
(functionally) related to a \coqe!Rows.T B!, as long as we can
(functionally) relate \coqe!A! and \coqe!B!. Put otherwise, we ask
that \coqe!Rows.T!, and similarly \coqe!Cols.T! and \coqe!Slice8.T!,
come equipped with their free theorems~\citep{wadler:free-theorem}.

%% * Objective: purely functional model
%% ** rule of the game: "no i, no j"

Having modeled the core data type, we now turn to modeling the cipher
as a purely functional program. To stay clear from the temptation of
array mismanagement, we will forbid ourselves from index arithmetic:
no indexing an array-like structure by \coqe!i+1! or \coqe!j-i! in
this paper!

\subsection{Key mixing}

%% * Generically over any Boolean algebra
%% ** obvious: `bool`
%% ** useful: `F bool`
%% *** for any commutative applicative functor F

Much like $\lambda$ in functional programming, the exclusive-or
features prominently in the cryptographic liturgy. In particular, it
is instrumental to mix the derived keys during the ciphertext
computation. For genericity, we model this layer as an operation
defined over any Boolean algebra~\citep{boolean-alg}:
%
\coqfrom{Skinny/Base}{addroundkey}

%% * Structure
%% ** `bool` is Boolean algebra
%% ** Rows.T, Cols.T and Slice8.T are commutative applicative functors

The type \coqe!bool! is an obvious instance of a Boolean algebra.
However, any commutative applicative~\citep{applicative-functor,
  commutative-appfunc} functor \coqe!F! applied to a Boolean algebra
also yields a Boolean algebra (dispatching the Boolean operations
pointwise to the underlying elements). For the sake of completeness,
we recall that an applicative functor offers the following operations:
%
\coqfrom{Algebra}{Applicative}

\subsection{Diffusion}

%% * Naperian functor: functional indexing
%% ** definition
%% ** also known as: representable functor

The role of the diffusion layer is to divert individual bits across the
whole structure. Inevitably, this calls for a notion of indexing.
Naperian functors~\citep{naperian} identify a type \coqe!Ix! of
indices as the logarithm of an exponential type:
%
\coqfrom{Algebra}{Naperian}
%
through the fact that \coqe!lookup! and \coqe!init! form a bijection.
This is also known as a representable functor in mathematical
circles~\citep{mclane:working-categorist}

%% ** example: Rows.T indexed by Rows.Ix
%% ** example: Cols.T indexed by Cols.Ix
%% ** example: Slice8.T indexed by Slice8.Ix

Following our earlier discussion, we ask that \coqe!Rows.T!
(respectively, \coqe!Cols.T!) is a Naperian functor indexed by a set
of 4 elements. Similarly, \coqe!Slice8.T! is a Naperian functor
indexed by 8 elements. We defined
%
\coqfrom{Rows}{Ix}
%
as the logarithm of \coqe!Rows.T!,
%
\coqfrom{Cols}{Ix}
%
as the logarithm of \coqe!Cols.T!, and
%
\coqfrom{Slice8}{Ix}
%
as the logarithm of \coqe!Slice8.T!.

Note that we have a notion of indexing but no arithmetic: we remain in
line with our objectives. Note also that being Naperian implies being
a commutative applicative functor. As a corollary, we get that any
combination of \coqe!Rows.T!, \coqe!Cols.T! and \coqe!Slice8.t!
applied to \coqe!bool! yields a Boolean algebra (over which we can
therefore write combinational circuits).

%% ** derived operation: Naperian implies commutative applicative
%% ** derived operation: structure of indices

Applying \coqe!init! to the identity function, we generically compute
the container of indices:
%
\coqfrom{Algebra}{indices}

We can witness the fact that \coqe!Rows.T! contains at least as many
elements as \coqe!Cols.T! through the following construction:
%
\coqfrom{Skinny/Base}{indices_C}

Note that, since they actually have the same number of elements, we
could also go the other way around, defining an inhabitant of
\coqe!Cols.T Rows.Ix!. We do not need this construction for Skinny. It
is necessary in the case of Gift, which proceeds by transposition of
$4 \times 4$ matrix.

%% * Foldable: left-to-right iteration
%% ** induce: left-to-right order
%% ** derived operation: left and right rotation

The first step of diffusion process consists in performing a right
rotation by a specific amount over each individual column. However, we
do not have enough structure to identify a ``right'' or ``left''
direction over our containers. The standard tool of the trade to
achieve this effect is the notion of foldable
functor~\citep{foldable}:
%
\coqfrom{Algebra}{Foldable}

Taking the monoid to be \coqe!list A! (under nil and concatenation,
\ie{} the free monoid construction), we can read off the
directionality we were after.
%
More usefully for our purposes, we can derive generic left and right
rotation rotation operators for any foldable functor \coqe!T! Naperian
over a type \coqe!Ix!:
%
\coqfrom{Rows}{ror_sig}
\coqfrom{Rows}{rol_sig}

\TODO{Foldable to ror/rol}

%% * Application: shiftows
%% ** figure: rotate each column by the the row index

After suitable generalization, this leads us to the following model
for the operation depicted in Figure~\ref{fig:shiftrows}
%
\coqfrom{Skinny/Base}{shiftrows}
%
which represents the first half of the diffusion layer.

%% * Application: mixcolumns
%% ** figure: psi_ >-> ror 1

The second half corresponds to a combinational circuit \coqe!psi_!
%
\coqfrom{Skinny/Base}{psi}
%
which exercises the Naperian structure of \coqe!Rows.T! together with
some underlying Boolean algebra. This is followed by a right rotation
of rows. Altogether, this defines the operation depicted in Figure~\ref{fig:mixcolumns}
%
\coqfrom{Skinny/Base}{mixcolumns}

\begin{figure}[htp]
  \caption{Shiftrows}
  \label{fig:shiftrows}

  \caption{Mixcolumns}
  \label{fig:mixcolumns}
\end{figure}

\subsection{Confusion}

% * Naperian functor
%% ** derived operation: permutation

The confusion layer builds upon two permutations over \coqe!Slice8.T!.
Once again, this is merely exploiting the Naperian structure of
\coqe!Slice8.T!: for any Naperian functor \coqe!F! indexed by
\coqe!Ix!, there exists a generic permutation operator
%
\coqe!perm : forall A, (Ix -> Ix) -> F A -> F A!.
%
This leads to the following definitions:
%
\coqfrom{Skinny/Base}{perms}

%% * Application: subcells

At the heart of this layer stands the substitution box \coqe!s8!,
which is just another combinational circuit (Figure~\ref{fig:s8}):
%
\coqfrom{Skinny/Base}{s8}

These building blocks are chained together to form the confusion
layer, \coqe!subcells!:
%
\coqfrom{Skinny/Base}{subcells}

\begin{figure}

  \caption{S8}
  \label{fig:s8}
\end{figure}

\subsection{Making the rounds}
\label{sec:assembling}

%% * state: natural composition order
%% ** -> thinking required
%% *** /> not necessarily efficient implementation
%% *** -> see sec:bitslicing

A round of Skinny is obtained by composing the following operations in
this particular order:
%
\begin{enumerate}
\item \coqe!subcells_ : forall {B : Type}, Boolean B -> Slice8.T B -> Slice8.T B!
\item \coqe!add_round_key_ : forall {B : Type}, Boolean B -> B -> B -> B!
\item \coqe!shiftrows_ : forall {A : Type}, Rows.T (Cols.T A -> Cols.T A)!
\item \coqe!mixcolumns_ : forall {B : Type}, Boolean B -> Rows.T B -> Rows.T B!
\end{enumerate}

Each of these operations will have to be lifted up to proceed over the
entire state of the cipher. We shall therefore settle on the most
convenient composition order for the type constructors \coqe!Rows.T!,
\coqe!Cols.T! and \coqe!Slice8.T!.

%% *** solve type constraints
%% **** shiftrows_ imposes on Rows.T . Cols.T
%% **** subcells_, add_round_key_ and mixcolumns_ naturally at `B = bool`
%% **** -> state

The solution can be read off from the type constraints induced by
individual elements and their composition. First, the type of
\coqe!shiftrows_! dictates that we compose \coqe!Rows.T! followed by
\coqe!Cols.T!. Then, we observe that \coqe!subcells_! can be grounded
to \coqe!bool! without disturbance. This leads to following 
composition order:
%
\coqfrom{Skinny/Spec}{state}

%% * round: chain components, lifting functorially as necessary
%% ** subcells_: map-ed
%% ** add_round_key: applied pointwise
%% ** shiftrows_: applied pointwise
%% ** mixcolumns_: over non trivial Boolean structure `Cols.T Slice8.T bool`

Consequently, our hands are forced when it comes to lifting up
individual components:
%
\coqfrom{Skinny/Spec}{round}
%
In particular, we have that \coqe!subcells_! is iterated pointwise
across rows and columns, \coqe!add_round_key_! handles the entire cube
as the support for a Boolean algebra, \coqe!shiftrows_! applicatively
applies its column transformation over individual rows and
\coqe!mixcolumns! handles horizontal faces of the cube as Boolean
algebras.

%% * primitive: merely iterate over key schedule
%% ** computed offline

The overall primitive is nothing but the iteration of a single round
over the list of round subkeys, produced by an offline key schedule
procedure:
%
\coqfrom{Skinny/Spec}{skinny}

\section{Bitslicing}
\label{sec:bitslicing}

%% * Cost analysis: memory usage
%% ** naïvely: 4*4*8 bytes (at least) storing a boolean each
%% ** -> waste of memory bandwidth
%% ** -> waste of register space

If we were to adopt \coqe!state! as our actual representation, the
memory usage of the resulting program would quite poor. It would take
at least 128 bytes to store that many bits (assuming that booleans are
stored in the smallest addressable memory size, a byte). This would be
a terrible waste of memory bandwidth and very poor register usage.

%% * Packed representation
%%   -> improve memory density
%%   /> promise: efficiently computable
%%    -> not quite true, see sec:fixslicing
%% ** state ~ Rows.T . Cols.T (Slice8.T bool) and code polymorphic
%% *** | Rows.T . Cols.T bool | is 16 bits
%% **** -> pack in register
%% *** |Slice8.T (Rows.T . Cols.T bool)| is 8 machine words
%% **** -> v. reasonable register pressure

We must therefore look toward adopting a \emph{packed} representation,
grouping individual booleans into a single machine word. Doing so
would improve memory density. However, we have to be very careful that
we can still efficiently compute over this packed representation. In
effect, for Skinny and Gift, this is not quite true: we will have to
wait until Section~\ref{sec:fixslicing}, which operates on the same
data representation, to have our cake and eat it too.

%% * Remark: AoS vs. SoA
%% ** /> bit level

Bitslicing is the cryptographer's jargon for such a representation
change. In the case of Skinny, we observe that only \coqe!shiftrows_!
imposes a strict composition order of \coqe!Rows.T! followed by
\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out the
composition. We would thus have the type \coqe!Rows.T (Cols.T bool)!
representing 16 bits and fitting snugly in a 32 bits machine word. The
overall composition \coqe!Slice8.T (Rows.T (Cols.T bool))! would take
8 registers in total, imposing a very reasonable amount of register
pressure to the CPU.
%
In effect, to return back to the jargon of compiler designers, we
propose to turn a structure-of-arrays (SoA) into an
array-of-structures (AoS), only we are dealing with bit-level
quantities here.

%% ** 32 bits registers: parallelism
%% *** <- embedded ARM Cortex platforms
%% **** not nostalgia for vinyls and vintage computing
%% *** -> Double.T is Naperian
%% **** -> Double.T bool is Boolean
%% *** -> 8 registers of 32 bits machine words

With climate change threatening the Planet, storing only 16 bits in a
32 bits word would be borderline unethical. We can further increase
register usage and instantly double throughput by processing two
blocks at the same time. To this end, we introduce a new Naperian
functor, \coqe!Double.T!, indexed by the two element set \coqe!Fst!
and \coqe!Snd!. This leads to
%
\coqfrom{Skinny/Bitslice}{state}
%
which denote our intent to manipulate objects of type \coqe!reg32! as
if they were stored in a single machine word. In the present work,
this observation will remain at the state of wishful thinking. One can
think of the code in this section (as well as the fixsliced code, in
the next section) as playing, themselves, the role of specifications
to lower-level refinements. Our previous work on Usuba suggests that
we will be able to deliver on this promise in the future. Further,
remark that the \coqe!state! now corresponds to two intermingled
blocks but as we shall process them in lock-step throughout the
cipher, this does not make much difference.

%% * Essence of bitslicing: packing + *efficient* on machine word
%% ** caveat: leaving out efficiency here
%% *** /> present work: stop at symbolic
%% *** cf. Usuba
%% ** non-example: app shiftrows is very costly here
%% *** cf. sec:fixslicing

%% * Naperian functor
%% ** derived op: reindex
%% *** -> operationally: change iteration order

Given two Naperian functors \coqe!F! and \coqe!G!, there exists a
generic notion of reindexing~\citep{gibbons:apl-icative}:
%
\coqfrom{Algebra}{reindex}

Operationally, reindexing effects a change of the iteration order of
the composed data container: we go from supporting iteration over
\coqe!G! within an external iteration over \coqe!F! to an interation
over \coqe!F! within an external iteration over \coqe!G!.

%% ** derived: systematic "orthogonalization" process
%% *** cite: Knuth

Using such reindexing, we can specify the transposition process
relating two input blocks (following the specification) and their
bitsliced counterpart:
%
\coqfrom{Skinny/BitsliceProof}{to_bitslice}

%% * Correctness
%% ** R: equivalent data up-to change of representation
%% ** must be preserved across the assembly line

The correctness of a candidate bitsliced implementation, dubbed
\coqe!Bitslice.skinny!, amounts to the following:
%
\begin{theo}[Correctness of bitslicing]
  For any list of pairs of round subkeys (of type
  %
  \coqe!list (Spec.state * Spec.state)!)
  %
  and for any pair of input blocks (of type
  %
  \coqe!list Spec.state!
  %
  individually), we have that a single execution of
  \coqe!Bitslice.skinny! produces the same output (after
  transposition) as two runs of the specification \coqe!Spec.skinny!:

  \coqfrom{Skinny/BitsliceProof}{theorem}
\end{theo}

%% * Implementation
%% ** Push `R` through every individual step
%% ** Read the solution from the proof obligations

In practice, it is useful to generalize this statement and, instead,
state that \coqe!Spec.skinny! and \coqe!Bitslice.skinny! must preserve
the following relation
%
\coqfrom{Skinny/BitsliceProof}{invariant_bitslice}

The relational invariant will then naturally and compositionally
percolate through the various layers of the ciphers. We can then read
off the bitsliced code from the bitsliced types and relational
invariant. A single round of the cipher thus becomes:
%
\coqfrom{Skinny/Bitslice}{round}

%% * Performance analysis on machine words
%% ** subcells_, add_round_key, psi_
%% *** boolean operations or, and, etc. become native logical ops on machine words
%% *** great!
%% ** shiftrows, Rows.ror: fiddly bit-level masking
%% *** -> next step: make it go away
%% *** (mind-blowing trick)

Considering (very carefully!) the instanciations of \coqe!subcells_!,
\coqe!add_round_key_! and \coqe!psi_! (which is part of
\coqe!mixcolumns_!), we check that these can be efficiently
implemented over machine words: they only involve boolean operations,
applied bitwise over \coqe!reg32!.

Unfortunately, \coqe!shiftrows_! and \coqe!Rows.ror! involve some very
fiddly and computationally costly bit twiddling. These amount to
performing bit-level permutations within a machine word. This is a
frequent pain point when bitslicing the diffusion layer of ciphers: it
is also true in the case of Gift and AES, for instance. In the next
section, we present this one weird trick to get permutations to nearly
vanish.

\section{Fixslicing}
\label{sec:fixslicing}

%% * Splitting the permutation
%% ** machine word supported then bit-level
%% ** push and accumulate bit-level across code

Studying the Gift cipher, \citet{adominicai} noticed that the
permutation layer can be decompose into, first, a transformation that
can be efficiently implemented within a register and, second, a
permutation that commutes with the other layers (confusion and key
mixing). Similarly, the diffusion layer of Skinny can be decomposed as
follows
%
\begin{prop}
  \coqfrom{Skinny/FixsliceProof}{split_phi_sigma}
\end{prop}
%
%% ** phi: costly permutation
%% *** -> not applied
%
where \coqe!phi_! is a permutation with inverse \coqe!inv_phi_! that
captures all the costly bit-twiddling operations
%
\coqfrom{Skinny/Base}{phi}
\coqfrom{Skinny/Base}{inv_phi}
%
%% ** sigma: efficient
%% *** <- xor: machine word xor
%% *** <- Cols.ror: machine word rotation
%% **** ARM: barrel shifter: xor + shift
%
while, in turn, \coqe!sigma_! admits an efficient implementation on
machine words
%
\coqfrom{Skinny/Base}{sigma}

Indeed, aside from the logical operations, the rotations are in fact
almost free thanks to the \textsc{Arm} Cortex barrel
shifters.\footnote{By this remark, we hope to have established our
credentials as budding cryptographers.}

%% * Propagating phi
%% ** 4 phi almost converges to identity

Being able to postpone \coqe!phi_! would do us no good if we were
nonetheless forced to run it anyway. The second, crucial observation
is that, after only 4 steps, \coqe!phi_! is almost an identity.

\begin{prop}
  For any state \coqe!rs: Rows.T (Cols.T A)!, we have:

  \coqfrom{Skinny/FixsliceProof}{tau_phi4}
  %
  where \coqe!tau_! is defined as
  %
  \coqfrom{Skinny/Base}{tau}
  %
  and can be implemented reasonnably efficiently.
\end{prop}

In typical mode of operation, Skinny will perform between 32 and 56
rounds. Being able to accumulate \coqe!phi_! across chunks of 4
iterations, we can effectively make them disappear, leaving only a
\coqe!tau_! step every four iterations.

%% ** phi commutes with subcells
%% *** <- subcells is applied pointwise

We verify that \coqe!phi_! commutes with \coqe!subcells_!:
%
\begin{prop}
  \coqfrom{Skinny/FixsliceProof}{comm_phi_subcells}
\end{prop}

This is intuitively obvious since \coqe!subcells_! proceeds pointwise
over rows and columns, hence changing their respective position does
not influence the outcome.

%% ** phi commutes with add_round_key
%% *** <- add_round_key is applied pointwise
%% *** remark: need to transform input key

Similarly and for the same reason, \coqe!phi_! commutes with
\coqe!add_round_key!:
%
\begin{prop}
  \coqfrom{Skinny/FixsliceProof}{comm_phi_add_round_key}
\end{prop}

Note that we have to transform the input key before hand, so as to
keep both structures aligned. In particular, this means that we do not
have to correct the first round subkey, we will have to apply
\coqe!inv_phi! to the second round subkey, \coqe!iter 2 inv_phi! to
the third round subkey and, finally, \coqe!iter 3 inv_phi! to the
fourth round subkey. The fifth round subkey starts back in sync,
\etc{}

%% ** permutation layer: emits a phi and produce a localized code
%% *** <- inv_phi_ . sigma_ . phi_ is stable

The crux of the matter is the interaction between \coqe!phi_! and the
diffusion layer. We observe that putting a state \coqe!phi_ s! into
the diffusion layer first yield a transformation relativized followed
by the emission of two \coqe!phi_! steps out of the layer:
%
\begin{coro}
  \coqfrom{Skinny/FixsliceProof}{absorb_phi_sigma1}
\end{coro}

%% *** -> generally: phi accumulates
%% **** define stable mixcolumn_mod
%% **** at 4 + tau: back to identity

This means that, in turn, the subsequent diffusion layer will have to
absorb two \coqe!phi_! steps. We thus generalize our statement for any
number of steps, the previous proposition being taken at $n = 1$.
There are 4 relativized diffusion layers, namely 
%
\coqe!mixcolumns_mod 0!,
%
\coqe!mixcolumns_mod 1!,
%
\coqe!mixcolumns_mod 2!
%
and
%
\coqe!mixcolumns_mod 3!
%
\coqfrom{Skinny/Fixslice}{mixcolumns_mod}
%
and they correspond to the diffusion layers absorbing that many
\coqe!phi_! steps:
%
\begin{prop}
  \coqfrom{Skinny/FixsliceProof}{absorb_phi_sigma}
\end{prop}

%% * Round
%% ** shiftrows spread out across 4 mod steps + tau
%% ** Reset by tau after 4

Given 4 (suitably generated) round subkeys, we merely have to chain 4
rounds with the corresponding diffusion layer, ending with \coqe!tau_!
to resynchronize back to identity
%
\coqfrom{Skinny/Fixslice}{round}

%% * Theorem
%% ** Proof sketch: maintain 4 relations
%% ** permbits moves from one to the next
%% ** fixsliced keys are organized in packs of 4
%% *** <- suitably pre-transformed with iter n inv_phi

The correctness statemement is unsurprising, modulo some light
bureaucracy to ensure that round subkeys have been suitably generated
%
\begin{theo}
  \coqfrom{Skinny/FixsliceProof}{theorem}
\end{theo}

The meat of the proof consists in, as for bitslicing, generalizing
this statement to a relational one. With fixslicing, there are in fact
4 relations, depending on the number of \coqe!phi_! steps accumulated.
We move from one to the next every time we go through a diffusion
layer, eventually resetting from 3 to 0.

%% * Actual implementation of mixcolumn_mod (1 + n)?

%% ** want: closed term in language of Boolean algebra (xor, and,
%%           ...), Naperian (init, lookup) and Foldable (ror, rol)
%% *** solution
%% **** make them Opaque
%% **** orient their equational theory
%% **** normalize `mixcolumns_mod 2`
%% **** autosubst
%% **** read off implementation from equations
%% ** what was done historically: propose a solution (and check correctness)

The definitions of \coqe!mixcolumn_mod! is somewhat disappointing: we
would be hard-pressed to make any precise claim about the potentiality
of an efficient implementation. The current form is symbolically
useful, as it allows us to easily the correctness of fixslicing.
However, it remains a mathematical specification, not a piece of
software. In effect, we are looking for an open term expressed in the
language of Boolean algebras (\coqe!xor!, \etc{}), Naperian operations
(\coqe!init!, \coqe!lookup!) and foldable operations (more precisely,
\coqe!ror! and \coqe!rol!).

One solution\footnote{Needless to say, we actually started off with
the answers, kindly provided by~\citet{adomi}, and crawled our way
back to see if an alternative approach would be feasible.}, which
consists in using Coq as a term rewriting engine, is to postulate the
existence of such an explicit form, let us call it
%
\coqe!mixcolumns_mod_explicit!.
%
We then claim that this definition ought to be equivalent to its
specification, \coqe!mixcolumns_mod! for $n \in \{0, 1, 2, 3\}$.
Making sure that the operations of Boolean algebra, Naperian and
foldable functors are opaque, $\beta$ reduction yields a term in our
target language. However, it suffers from significant redundancy: by
working over the term algebra, we have not quotiented out the
equational theory. To do so, we orient the equational theory in a
rewrite database and apply the \coqe!autosubst! tactics to obtain
simplified forms. Similarly, we retrieve convenient \coqe!let! forms
thanks to the \coqe!set (ident := term)! tactics. We are then left
with reading off the actual definition of
\coqe!mixcolumns_mod_explicit!.

\section{Conclusion}

We have thus completed our journey toward a rationalized treatment of
bitslicing and fixslicing in their quintessential form, taking the
Skinny cipher as our running example.
%
%% * Other applicative: Gift
%% ** Specification: only thinking: Rows.T Cols.T Slice4.T bool
%% *** rest is rote functionalisation
%% ** Bitslicing: extract Slice4.T out and Double
%% ** Fixslicing: decompose permutation into in-register followed by 90 degree rotation
%% *** cancel out after 4
%% *** -> simpler
%
We have similarly treated the Gift cipher, whose design is in fact at
the origin of the fixslicing technique~\citep{adomi}. Gift is
specified in terms of the composition
%
\coqe!Rows.T (Cols.T (Slice4.T bool))!
%
where \coqe!Slice4.T! denotes a depth of 4 slices. In bitsliced form,
we once again extract out the \coqe!Slice4.T! functor and
\coqe!Double.T! the amount of data processed per run. Fixslicing is
conceptually easier to understand as the diffusion layer is defined as
the combination of an in-register transformation followed by a
90-degree rotation of the matrix of rows and columns. It simply
cancels out after 4 steps.

%% * Dirty secret
%% ** No proof! vm_compute; reflexivity

Of note, the propositions that lead to the bitslicing and fixslicing
correctness theorems hide a somewhat surprising secret: all the proofs
were obtained by \coqe!vm_compute; congruence!. Having defined
\coqe!Rows.T!, \coqe!Cols.T!, \coqe!Slice4.T! and \coqe!Slice8.T! as
records with primitive projections, all the identities boil down to
the specification, bitsliced forms and fixsliced forms to have the
same $\beta$-normal $\eta$-long form through the suitable equivalence
relation!

%% * Future work
%% ** Bridge to low-level implementation
%% *** reg32 fits in machine word
%% *** exploit parametricity: representation change
%% **** clear blueprint for implem and proofs
%% ** Usuba 2.0: beyond bitslicing
%% *** Derive fixsliced permutations
%% *** Check commutativity and convergence

%% ** Litmus test: automatically fixsliced AES
%%   -> stay tuned

As part of future work, we intend to extend our formalization to
bridge the gap to machine word. Once again, we expect parametricity to
kick in: we currently have an implementation defined over \coqe!reg32!
(and its supporting operations) which we intend to be equivalent to an
implementation defined over the type of 32 bits machine words (and its
supporting operations).

We also wish to pursue the automated generation of fixsliced code,
eventually implementing a proper simplification engine. In the
process, we ought to develop methods to check for commutativity of the
diffusion layer.

A litmus test to this project will be our ability to deliver a
fixsliced implementation of AES: mark our machine words!

\bibliographystyle{alpha-fr}
\bibliography{paper}

\end{document}

%% Local Variables:
%% mode: latex
%% mode: outline-minor
%% outline-regexp: "%% [*\f]+"
%% outline-level: outline-level
%% End:

### 2024-10-12 paper.tex
Add figures

@@
 \usepackage{enumitem}
+\usepackage{subcaption}
 
@@
   compositional structure at the type-level, which enables systematic
-  code transformation, such as bitslicing and fixslicing. Our work
-  rests on a Coq development providing the specification of two
-  ciphers, Skinny and Gift, deriving a bitsliced and fixsliced
-  implementation for each and proving their correctness with respect
-  to their specifications.
+  code transformation, known in the cryptographical folklore as
+  ``bitslicing'' and ``fixslicing''. Our work rests on a Coq
+  development providing the specification of two ciphers, Skinny and
+  Gift, deriving a bitsliced and fixsliced implementation for each and
+  proving their correctness with respect to their specifications.
 
@@
 cipher~\citep{gift-spec, admon:gift}. Both ciphers are provided in the
-aaccompanying source code~\footnote{\url{TODO}}. For clarity, we shall
+aaccompanying source code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity, we shall
 focus exclusively on Skinny as our running example here.
@@
 \begin{figure}
-  %\includegraphics{skinny_state}
+  \begin{center}
+    \includegraphics[draft=false,height=5cm]{skinny_state}
+  \end{center}
   \caption{Skinny state matrix}
@@
 \begin{figure}[htp]
-  \caption{Shiftrows}
+  \centering
+  \begin{subfigure}[t]{0.45\textwidth}
+    \centering
+    \includegraphics[draft=false,height=2.8cm]{shiftrows}
+    \caption{Shiftrows (from~\citet{skinny-spec})}
     \label{fig:shiftrows}
-
-  \caption{Mixcolumns}
+  \end{subfigure}
+  \begin{subfigure}[t]{0.45\textwidth}
+    \centering
+    \includegraphics[draft=false,height=2.8cm]{mixcolumns}
+    \caption{Mixcolumns (from~\citet{skinny-spec})}
     \label{fig:mixcolumns}
+  \end{subfigure}
 \end{figure}
@@
 \begin{figure}
-
-  \caption{S8}
+  \centering
+  \includegraphics[draft=false,height=5cm]{s8}
+  \caption{Substitution box S8 (from~\citet{skinny-spec})}
   \label{fig:s8}

### 2024-10-12 paper.tex
Add refs

@@
 implementation of symmetric ciphers. This has been slowly gnawing at
-compiler writers, for whom job security~\citep{bernstein} has become a
-major concern. Luckily for us, every now and then, someone
-inadvertently let the genies out of their Klein bottle and we get to
-read about the subtle art of ``bitslicing''~\citep{kwan:bitslicing}
+compiler writers, for whom job
+security~\citep{bernstein:death-compiler,venkatesh:compiler-security}
+has become a major concern. Luckily for us, every now and then,
+someone inadvertently let the genies out of their Klein bottle and we
+get to read about the subtle art of
+``bitslicing''~\citep{biham:DES} and
+``fixslicing''~\citep{adominicai:fixslicing-gift,
+  adomnicai:fixslicing-AES-like}.
 
@@
 optimized code, starting from an high-level description of the cipher.
-This is already the case for bitslicing~\citep{usuba}, adding support
+This is already the case for bitslicing~\citep{mercadier:PhD}, adding support
 for fixslicing is next on the list.
@@
 To avoid side-channel attack (for example, based on
-timing~\citep{time-side-channel}), this pipeline can be more precisely
-understood as a combinational circuit. We are forbidden from
+timing~\citep{bernstein:timing-attack-AES}), this pipeline can be more
+precisely understood as a combinational circuit. We are forbidden from
 performing control-flow operations on secret data, leaving only
@@
 Following the NSA involvement in the design of the DES diffusion
-layer~\citep{des-history}, cryptographers have been advocating for
-``nothing-up-my-sleeve'' designs (which does not make magic tricks
-impossible, just harder~\citep{bernstein:something-up-my-sleeves}).
-For example, the diffusion layer of the AES standard is based on the
-usual multiplicative inverse over the finite field induced by an
+layer~\citep{johnson:cold-war-crypto}, cryptographers have been
+advocating for ``nothing-up-my-sleeve'' designs (which does not make
+magic tricks impossible, just
+harder~\citep{bernstein:something-up-my-sleeves}). For example, the
+diffusion layer of the AES standard is based on the usual
+multiplicative inverse over the finite field induced by an
 innocent-looking polynomial. To support such design, the state of
@@
 efficiently on software. The era of hardware-based cryptography is
-long gone, even for embedded platforms~\citep{lightweight-crypto}. To
-achieve high-throughput, ciphers are typically designed so as to be
-admit a bitsliced implementation. Bitslicing enables a form of
-parallel execution \emph{within} machine words. We thus gain access to
-a form of CPU-level ``scale out'' to larger machine words, in
-particular through single-instruction multiple-data (SIMD)
-instructions. Doubling the size of machine words yields nearly twice
-the throughput, without any effort. Evidences tend to suggest that
-cryptographers obtain bitsliced designs through sheer intellectual
-might. It is in fact the sole purpose of the ``Section 5'' scene to
-exhibit a bitsliced witness in all the gory details.
+long gone, even for embedded
+platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
+high-throughput, ciphers are typically designed so as to be admit a
+bitsliced implementation. Bitslicing enables a form of parallel
+execution \emph{within} machine words. We thus gain access to a form
+of CPU-level ``scale out'' to larger machine words, in particular
+through single-instruction multiple-data (SIMD) instructions. Doubling
+the size of machine words yields nearly twice the throughput, without
+any effort. Evidences tend to suggest that cryptographers obtain
+bitsliced designs through sheer intellectual might. It is in fact the
+sole purpose of the ``Section 5'' scene to exhibit a bitsliced witness
+in all the gory details.
 
@@
 the wake of the Squiggol school of
-programming~\citep{wg21,gibbons:naperian}. Because our
+programming~\citep{bird:squiggol,roconnor:functor-oriented}. Because our
 government-approved programming language does not yet support ad-hoc
-polymorphism~\citep{remy:explicit-poly}, the French community may be
+polymorphism~\citep{white:modular-implicit}, the French community may be
 missing out on some interesting programming patterns. The present
@@
 touching upon implementation aspects (which remain the raison d'être
-of the project!). Our experience designing the Usuba~\citep{usuba}
+of the project!). Our experience designing the Usuba~\citep{mercadier:PhD}
 programmign language and implementing its compiler gives us some
@@
 went from there down to deeply-embedded assembly
-code~\citep{chlipala}. Besides the ability to prove the correctness
-and, even more usefully, test our code, this also enables us to easily
-prototype a fixslicing compiler, using Coq's \texttt{autorewrite}
-tactics to emulate a rule-based simplification
-engine~\citep{haskell:simplier} (part of an imaginary optimizing
+code~\citep{chlipala:coq-crypto}. Besides the ability to prove the
+correctness and, even more usefully, test our code, this also enables
+us to easily prototype a fixslicing compiler, using Coq's
+\texttt{autorewrite} tactics to emulate a transformation-based simplification
+engine~\citep{peyton-jones:haskell-optim} (part of an imaginary optimizing
 compiler).
@@
 on the specification, bitslicing and fixslicing of the Skinny
-cipher~\citep{skinny-spec, adomin:skinny}. The latter worked on the
-specification, bitslicing and fixslicing of the Gift
-cipher~\citep{gift-spec, admon:gift}. Both ciphers are provided in the
-aaccompanying source code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity, we shall
-focus exclusively on Skinny as our running example here.
+cipher~\citep{beierle:skinny, adomnicai:fixslicing-AES-like}. The
+latter worked on the specification, bitslicing and fixslicing of the
+Gift cipher~\citep{subhadeep:gift, adominicai:fixslicing-gift}. Both
+ciphers are provided in the aaccompanying source
+code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
+we shall focus exclusively on Skinny as our running example here.
 
@@
 that \coqe!Rows.T!, and similarly \coqe!Cols.T! and \coqe!Slice8.T!,
-come equipped with their free theorems~\citep{wadler:free-theorem}.
+come equipped with their free theorems~\citep{reynolds:parametricity,
+  wadler:free-theorem}.
 
@@
 computation. For genericity, we model this layer as an operation
-defined over any Boolean algebra~\citep{boolean-alg}:
+defined over any Boolean algebra~\citep{givant:boolean-algebras}:
 %
@@
 The type \coqe!bool! is an obvious instance of a Boolean algebra.
-However, any commutative applicative~\citep{applicative-functor,
-  commutative-appfunc} functor \coqe!F! applied to a Boolean algebra
-also yields a Boolean algebra (dispatching the Boolean operations
-pointwise to the underlying elements). For the sake of completeness,
-we recall that an applicative functor offers the following operations:
+However, any commutative applicative~\citep{mcbride:applicative}
+functor \coqe!F! applied to a Boolean algebra also yields a Boolean
+algebra (dispatching the Boolean operations pointwise to the
+underlying elements). For the sake of completeness, we recall that an
+applicative functor offers the following operations:
 %
@@
 whole structure. Inevitably, this calls for a notion of indexing.
-Naperian functors~\citep{naperian} identify a type \coqe!Ix! of
+Naperian functors~\citep{hancock:napierian} identify a type \coqe!Ix! of
 indices as the logarithm of an exponential type:
@@
 This is also known as a representable functor in mathematical
-circles~\citep{mclane:working-categorist}
+circles~\citep{maclane:working-categorist}
 
@@
 achieve this effect is the notion of foldable
-functor~\citep{foldable}:
+functor~\citep{yorgey:typeclassopedia}:
 %
@@
     \includegraphics[draft=false,height=2.8cm]{shiftrows}
-    \caption{Shiftrows (from~\citet{skinny-spec})}
+    \caption{Shiftrows (from~\citet{beierle:skinny})}
     \label{fig:shiftrows}
@@
     \includegraphics[draft=false,height=2.8cm]{mixcolumns}
-    \caption{Mixcolumns (from~\citet{skinny-spec})}
+    \caption{Mixcolumns (from~\citet{beierle:skinny})}
     \label{fig:mixcolumns}
@@
   \includegraphics[draft=false,height=5cm]{s8}
-  \caption{Substitution box S8 (from~\citet{skinny-spec})}
+  \caption{Substitution box S8 (from~\citet{beierle:skinny})}
   \label{fig:s8}
@@
 Given two Naperian functors \coqe!F! and \coqe!G!, there exists a
-generic notion of reindexing~\citep{gibbons:apl-icative}:
+generic notion of reindexing~\citep{gibbons:apl}:
 %
@@
 
-Studying the Gift cipher, \citet{adominicai} noticed that the
+Studying the Gift cipher, \citet{adominicai:fixslicing-gift} noticed that the
 permutation layer can be decompose into, first, a transformation that
@@
 One solution\footnote{Needless to say, we actually started off with
-the answers, kindly provided by~\citet{adomi}, and crawled our way
+the answers, kindly provided by~\citet{adominicai:fixslicing-gift}, and crawled our way
 back to see if an alternative approach would be feasible.}, which
@@
 We have similarly treated the Gift cipher, whose design is in fact at
-the origin of the fixslicing technique~\citep{adomi}. Gift is
-specified in terms of the composition
+the origin of the fixslicing
+technique~\citep{adominicai:fixslicing-gift}. Gift is specified in
+terms of the composition
 %
@@
 
-\bibliographystyle{alpha-fr}
+\bibliographystyle{unsrtnat}
 \bibliography{paper}

### 2024-10-12 paper.tex
Ispell

@@
   compositional structure at the type-level, which enables systematic
-  code transformation, known in the cryptographical folklore as
+  code transformation, known in the cryptographic folklore as
   ``bitslicing'' and ``fixslicing''. Our work rests on a Coq
@@
 innocent-looking polynomial. To support such design, the state of
-modern ciphers is thus equipped with some matricial structure
-(oftentimes, manipulated as a 2D or 3D array of bits): the cipher is
-not specified over a flat sequence of individual bits but over a
-n-dimensional array of bits.
+modern ciphers operate over a certain matrix (oftentimes, a 2D or 3D
+array of bits): the cipher is not specified over a flat sequence of
+individual bits but over a n-dimensional array of bits.
 
@@
 of the project!). Our experience designing the Usuba~\citep{mercadier:PhD}
-programmign language and implementing its compiler gives us some
+programming language and implementing its compiler gives us some
 confidence that a syntax can indeed be tailored to suit this
@@
 been a time where we do not have to worry about termination, this is
-it! So we have implemented cryptographical primitive in pure Gallina.
+it! So we have implemented cryptographic primitive in pure Gallina.
 This is nothing to worry about: after all, more ambitious colleagues
@@
 Gift cipher~\citep{subhadeep:gift, adominicai:fixslicing-gift}. Both
-ciphers are provided in the aaccompanying source
+ciphers are provided in the accompanying source
 code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
@@
 the composed data container: we go from supporting iteration over
-\coqe!G! within an external iteration over \coqe!F! to an interation
+\coqe!G! within an external iteration over \coqe!F! to an iteration
 over \coqe!F! within an external iteration over \coqe!G!.
@@
 
-Considering (very carefully!) the instanciations of \coqe!subcells_!,
+Considering (very carefully!) the instances of \coqe!subcells_!,
 \coqe!add_round_key_! and \coqe!psi_! (which is part of
@@
 implemented over machine words: they only involve boolean operations,
-applied bitwise over \coqe!reg32!.
+applied bit-wise over \coqe!reg32!.
 
@@
 almost free thanks to the \textsc{Arm} Cortex barrel
-shifters.\footnote{By this remark, we hope to have established our
+shifter.\footnote{By this remark, we hope to have established our
 credentials as budding cryptographers.}
@@
   %
-  and can be implemented reasonnably efficiently.
+  and can be implemented reasonably efficiently.
 \end{prop}
@@
 rounds with the corresponding diffusion layer, ending with \coqe!tau_!
-to resynchronize back to identity
+to re-synchronize back to identity
 %
@@
 
-The correctness statemement is unsurprising, modulo some light
+The correctness statement is unsurprising, modulo some light
 bureaucracy to ensure that round subkeys have been suitably generated

### 2024-10-12 paper.tex
Rephrase

@@
   functors). We reveal that these ciphers exhibit some interesting
-  compositional structure at the type-level, which enables systematic
+  compositional structure at the type-level. This enables systematic
   code transformation, known in the cryptographic folklore as

### 2024-10-12 paper.tex
Simplify

@@
 goes that, behind these padlocked doors, they have developed a set of
-domain-specific techniques to deliver very high-throughput software
+domain-specific techniques to deliver high-throughput software
 implementation of symmetric ciphers. This has been slowly gnawing at
@@
 someone inadvertently let the genies out of their Klein bottle and we
-get to read about the subtle art of
-``bitslicing''~\citep{biham:DES} and
-``fixslicing''~\citep{adominicai:fixslicing-gift,
+get to read about the subtle art of ``bitslicing''~\citep{biham:DES}
+and ``fixslicing''~\citep{adominicai:fixslicing-gift,
   adomnicai:fixslicing-AES-like}.

### 2024-10-12 paper.tex
Rephrase

@@
 implementation of symmetric ciphers. This has been slowly gnawing at
-compiler writers, for whom job
-security~\citep{bernstein:death-compiler,venkatesh:compiler-security}
-has become a major concern. Luckily for us, every now and then,
-someone inadvertently let the genies out of their Klein bottle and we
-get to read about the subtle art of ``bitslicing''~\citep{biham:DES}
-and ``fixslicing''~\citep{adominicai:fixslicing-gift,
+compiler writers, whose job security is now
+threatened~\citep{bernstein:death-compiler,venkatesh:compiler-security}.
+Luckily for us, every now and then, someone inadvertently let the
+genies out of their Klein bottle and we get to read about the subtle
+art of ``bitslicing''~\citep{biham:DES} and
+``fixslicing''~\citep{adominicai:fixslicing-gift,
   adomnicai:fixslicing-AES-like}.

### 2024-10-12 paper.tex
Grammar

@@
 
-The present work aims to cast these two notions in the language of
+The present work aims at casting these two notions in the language of
 functional programming. We shall see that bitslicing corresponds to a
@@
 to systematically transform the \emph{code} operating on such
-representation relies on suitably polymorphic definitions:
+representation rests on suitably polymorphic definitions:
 parametricity at its best! Fixslicing, on the other hand, will be

### 2024-10-12 paper.tex
Rephrase

@@
 rationalized through equational reasoning over purely functional
-terms. Identifying commuting terms will be of key importance there:
+terms. Identifying commuting expressions is of key importance there:
 programming with algebraic structures (rather than untamed arrays of

### 2024-10-12 paper.tex
Rephrase

@@
 primitive can be understood as a purely functional programs. It takes
-as input two blocks of data, one representing the input cleartext and
-the other representing the input keys (derived from a single key
-through a non-performance critical process called ``key schedule'',
-which we shall ignore here). The block size is of order of a hundred
-bits (typically, 64, 128 or 256 bits). The output is a block of data
-of the same size, the ciphertext.
+two inputs, the input ``cleartext'' and a list of ``round subkeys''
+(which are derived from a single key through a non-performance
+critical process called ``key schedule'' that we shall ignore here).
+The output is the ``ciphertext''. The cleartext, the ciphertext and
+each individual round subkeys are processed as ``blocks'' of binary
+data. The block size is of order of a hundred bits (typically, 64, 128
+or 256 bits).
 
@@
 all along: ciphers are described as operating on a ``state'',
-initialized to be the cleartext and read off as the ciphertext at the
-end. This state is updated by a ``round'' function that is repeated a
-given number of times (of the order of ten iterations). The exact
-number of iterations is a trade-off between latency (less iterations
-means faster computation) and security (more iterations makes for more
-costly cryptanalysis effort). The round function itself is built
-compositionally from 3 components: a key mixing layer (making the
-encryption process reversible), a confusion layer (smoothing local
-correlations in the inputs) and a diffusion layer (spreading
+initialized to be the cleartext and read off at the end as the
+ciphertext. This state is updated by a ``round'' function that is
+repeated a given number of times (of the order of ten iterations). The
+exact number of iterations is a trade-off between latency (less
+iterations means faster computation) and security (more iterations
+makes for more costly cryptanalysis effort). The round function itself
+is built compositionally from 3 components: a key mixing layer (making
+the encryption process reversible), a confusion layer (smoothing local
+correlations among the inputs) and a diffusion layer (spreading
 correlations across the whole output block).
@@
 To avoid side-channel attack (for example, based on
-timing~\citep{bernstein:timing-attack-AES}), this pipeline can be more
-precisely understood as a combinational circuit. We are forbidden from
-performing control-flow operations on secret data, leaving only
+timing~\citep{bernstein:timing-attack-AES}) over software
+implementations, this pipeline is implemented as a purely
+combinational circuit. In particular, we are forbidden from performing
+control-flow operations on secret data, leaving only
 statically-bounded loops. We are also forbidden from performing memory
-access based on the data flow, leaving only register spilling.
+access based on the secret data, leaving only register spilling.
 
@@
 innocent-looking polynomial. To support such design, the state of
-modern ciphers operate over a certain matrix (oftentimes, a 2D or 3D
-array of bits): the cipher is not specified over a flat sequence of
-individual bits but over a n-dimensional array of bits.
+modern ciphers operate is taken to represent a certain matrix
+(oftentimes, a 2D or 3D array of bits): the cipher is not specified
+over a flat sequence of individual bits but over a n-dimensional array
+of bits.
 
@@
 platforms~\citep{buchanan:nist-lightweight-crypto}. To achieve
-high-throughput, ciphers are typically designed so as to be admit a
+high-throughput, ciphers are typically designed so as to admit a
 bitsliced implementation. Bitslicing enables a form of parallel
@@
 bitsliced designs through sheer intellectual might. It is in fact the
-sole purpose of the ``Section 5'' scene to exhibit a bitsliced witness
-in all the gory details.
+sole purpose of the traditional ``Section 5'' scene to exhibit a
+bitsliced witness in all the gory details.
 
@@
 What would it take to turn this art form into an engineering
-principle? We answer through 3 contributions:
+principle? To answer, we make the following contributions:
 %
@@
 \item we identify the algebraic structure necessary to model the state
-  of a modern symmetric cryptography primitive
+  of modern symmetric cryptography primitives
   (Section~\ref{sec:spec}). Doing so, we delineate a mathematical
@@
   reasoning. To support such reasoning, we crucially rely on the
-  algebraic structure set forth in the Section~\ref{sec:spec}.
+  algebraic structure set forth in Section~\ref{sec:spec}.
 \end{itemize}

### 2024-10-12 paper.tex
Rephrase

@@
 stubbornly French functional programmers. The present work surfs on
-the wake of the Squiggol school of
-programming~\citep{bird:squiggol,roconnor:functor-oriented}. Because our
-government-approved programming language does not yet support ad-hoc
-polymorphism~\citep{white:modular-implicit}, the French community may be
-missing out on some interesting programming patterns. The present
-article can thus be read as a case-study in ``functor-oriented
-programming'', where the word ``functor'' is taken in his categorical
-meaning (recalled in Section~\ref{sec:spec} and unrelated to the one
-mentioned in the ML module system). In effect, we forbid ourselves
-from programming over a concrete datastructure: instead, we rely on
-(categorical) functors, peppered with some more structure, so as to 1.
-unlock algebraic manipulation during compilation and 2. exploit
-parametricity for reasoning.
+the wake of the Squiggol school of programming~\citep{bird:squiggol}.
+Because our government-approved programming language does not yet
+support ad-hoc polymorphism~\citep{white:modular-implicit}, the French
+community may be missing out on some interesting programming patterns.
+The present article can thus be read as a case-study in
+``functor-oriented programming''~\citep{roconnor:functor-oriented},
+where the notion of functor comes from category theory (recalled in
+Section~\ref{sec:spec} and unrelated to the notion used in ML module
+systems). In effect, we forbid ourselves from programming over a
+concrete datastructure: instead, we rely on (categorical) functors,
+peppered with some more structure, so as to 1. unlock algebraic
+manipulation during compilation and 2. exploit parametricity for
+reasoning.
 

### 2024-10-12 paper.tex
Rephrase

@@
 programming language and implementing its compiler gives us some
-confidence that a syntax can indeed be tailored to suit this
+confidence that a syntax can indeed be tailored to dress up this
 semantics.
@@
 us to easily prototype a fixslicing compiler, using Coq's
-\texttt{autorewrite} tactics to emulate a transformation-based simplification
-engine~\citep{peyton-jones:haskell-optim} (part of an imaginary optimizing
-compiler).
+\texttt{autorewrite} tactics to emulate a transformation-based
+simplification engine~\citep{peyton-jones:haskell-optim} (as part of
+an imaginary optimizing compiler).
 
@@
 ciphers are provided in the accompanying source
-code~\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
+code\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
 we shall focus exclusively on Skinny as our running example here.
@@
 code seem somewhat trivial, one should bear in mind that this
-simplicity was hard-won. All the more reason to, another day, design a
-programming language, so as to automatically inhabit this semantics,
-and write its compiler, so as to automatically optimize code following
-our theorems.
+simplicity was hard-won. All the more reason to, some other day,
+design a programming language, so as to automatically inhabit this
+semantics, and write its compiler, so as to automatically optimize
+code following our theorems.
 

### 2024-10-12 paper.tex
Rephrase

@@
 \newcommand{\etc}{\textit{etc.}}
+
+\newcommand{\Skinny}{\textsc{Skinny}}
+\newcommand{\Gift}{\textsc{Gift}}
+\newcommand{\AES}{\textsc{Aes}}
+\newcommand{\DES}{\textsc{Des}}
 
@@
   ``bitslicing'' and ``fixslicing''. Our work rests on a Coq
-  development providing the specification of two ciphers, Skinny and
-  Gift, deriving a bitsliced and fixsliced implementation for each and
+  development providing the specification of two ciphers, \Skinny{} and
+  \Gift{}, deriving a bitsliced and fixsliced implementation for each and
   proving their correctness with respect to their specifications.
@@
 
-Following the NSA involvement in the design of the DES diffusion
+Following the NSA involvement in the design of the \DES{} diffusion
 layer~\citep{johnson:cold-war-crypto}, cryptographers have been
@@
 harder~\citep{bernstein:something-up-my-sleeves}). For example, the
-diffusion layer of the AES standard is based on the usual
+diffusion layer of the \AES{} standard is based on the usual
 multiplicative inverse over the finite field induced by an
@@
 author and the Master thesis of the second author. The former worked
-on the specification, bitslicing and fixslicing of the Skinny
+on the specification, bitslicing and fixslicing of the \Skinny{}
 cipher~\citep{beierle:skinny, adomnicai:fixslicing-AES-like}. The
 latter worked on the specification, bitslicing and fixslicing of the
-Gift cipher~\citep{subhadeep:gift, adominicai:fixslicing-gift}. Both
+\Gift{} cipher~\citep{subhadeep:gift, adominicai:fixslicing-gift}. Both
 ciphers are provided in the accompanying source
 code\footnote{\url{https://github.com/pedagand/bitfix}}. For clarity,
-we shall focus exclusively on Skinny as our running example here.
+we shall focus exclusively on \Skinny{} as our running example here.
 
@@
 
-In this section, we intend to give a specification for the cipher
-Skinny, with an eye towards implementation. We shall therefore aim for
+In this section, we intend to give a specification for the \Skinny{}
+cipher, with an eye towards implementation. We shall therefore aim for
 a rather operational description, to give a sense of the computational
@@
 focal point is the \emph{state} of the cipher, which is specified as a
-3-dimensional array (Figure~\ref{fig:state}). We model each dimension
+3-dimensional matrix (Figure~\ref{fig:state}). We model each dimension
 in turn through a dedicated type constructor:
@@
 %
-or any other composition of these 3 type formers. They are all
+or any other composition of these 3 type constructors. They are all
 isomorphic to a sequence of 128 bits. We shall wait until
@@
   \end{center}
-  \caption{Skinny state matrix}
+  \caption{\Skinny{} state matrix}
   \label{fig:state}
@@
 could also go the other way around, defining an inhabitant of
-\coqe!Cols.T Rows.Ix!. We do not need this construction for Skinny. It
-is necessary in the case of Gift, which proceeds by transposition of
+\coqe!Cols.T Rows.Ix!. We do not need this construction for \Skinny{}. It
+is necessary in the case of \Gift{}, which proceeds by transposition of
 $4 \times 4$ matrix.
@@
 
-A round of Skinny is obtained by composing the following operations in
+A round of \Skinny{} is obtained by composing the following operations in
 this particular order:
@@
 we can still efficiently compute over this packed representation. In
-effect, for Skinny and Gift, this is not quite true: we will have to
+effect, for \Skinny{} and \Gift{}, this is not quite true: we will have to
 wait until Section~\ref{sec:fixslicing}, which operates on the same
@@
 Bitslicing is the cryptographer's jargon for such a representation
-change. In the case of Skinny, we observe that only \coqe!shiftrows_!
+change. In the case of \Skinny{}, we observe that only \coqe!shiftrows_!
 imposes a strict composition order of \coqe!Rows.T! followed by
@@
 frequent pain point when bitslicing the diffusion layer of ciphers: it
-is also true in the case of Gift and AES, for instance. In the next
+is also true in the case of \Gift{} and \AES{}, for instance. In the next
 section, we present this one weird trick to get permutations to nearly
@@
 
-Studying the Gift cipher, \citet{adominicai:fixslicing-gift} noticed that the
+Studying the \Gift{} cipher, \citet{adominicai:fixslicing-gift} noticed that the
 permutation layer can be decompose into, first, a transformation that
@@
 permutation that commutes with the other layers (confusion and key
-mixing). Similarly, the diffusion layer of Skinny can be decomposed as
+mixing). Similarly, the diffusion layer of \Skinny{} can be decomposed as
 follows
@@
 
-In typical mode of operation, Skinny will perform between 32 and 56
+In typical mode of operation, \Skinny{} will perform between 32 and 56
 rounds. Being able to accumulate \coqe!phi_! across chunks of 4
@@
 bitslicing and fixslicing in their quintessential form, taking the
-Skinny cipher as our running example.
+\Skinny{} cipher as our running example.
 %
@@
 %
-We have similarly treated the Gift cipher, whose design is in fact at
+We have similarly treated the \Gift{} cipher, whose design is in fact at
 the origin of the fixslicing
-technique~\citep{adominicai:fixslicing-gift}. Gift is specified in
+technique~\citep{adominicai:fixslicing-gift}. \Gift{} is specified in
 terms of the composition
@@
 A litmus test to this project will be our ability to deliver a
-fixsliced implementation of AES: mark our machine words!
+fixsliced implementation of \AES{}: mark our machine words!
 

### 2024-10-12 paper.tex
Rephrase

@@
 systems). In effect, we forbid ourselves from programming over a
-concrete datastructure: instead, we rely on (categorical) functors,
+concrete data structure: instead, we rely on (categorical) functors,
 peppered with some more structure, so as to 1. unlock algebraic
@@
 representation for our specification effort. We revisit this choice in
-Section~\ref{sec:bitslicing} when it comes to producing a
+Section~\ref{sec:bitslicing} when we are concerned with producing a
 memory-efficient representation.
@@
 
-Having modeled the core data type, we now turn to modeling the cipher
+Having modeled the core data types, we now turn to modeling the cipher
 as a purely functional program. To stay clear from the temptation of
-array mismanagement, we will forbid ourselves from index arithmetic:
-no indexing an array-like structure by \coqe!i+1! or \coqe!j-i! in
-this paper!
+array mismanagement, we disallow index arithmetic altogether: no
+indexing an array-like structure by \coqe!i+1! or, even worse,
+\coqe!j-i! in this paper!
 
@@
 Much like $\lambda$ in functional programming, the exclusive-or
-features prominently in the cryptographic liturgy. In particular, it
-is instrumental to mix the derived keys during the ciphertext
-computation. For genericity, we model this layer as an operation
-defined over any Boolean algebra~\citep{givant:boolean-algebras}:
+$\oplus$ features prominently in the cryptographic liturgy. In
+particular, it is instrumental to mix the derived keys during
+ciphertext computation. For genericity, we model this layer as an
+operation defined over any Boolean
+algebra~\citep{givant:boolean-algebras}:
 %
@@
 through the fact that \coqe!lookup! and \coqe!init! form a bijection.
-This is also known as a representable functor in mathematical
-circles~\citep{maclane:working-categorist}
+This is also known as a representable functor in categorical
+circles~\citep{maclane:working-categorist}.
 
@@
 
-Following our earlier discussion, we ask that \coqe!Rows.T!
-(respectively, \coqe!Cols.T!) is a Naperian functor indexed by a set
-of 4 elements. Similarly, \coqe!Slice8.T! is a Naperian functor
-indexed by 8 elements. We defined
+Following our earlier discussion, we ask for \coqe!Rows.T!
+(respectively, \coqe!Cols.T!) to be a Naperian functor indexed by a
+set of 4 elements. Similarly, \coqe!Slice8.T! must be a Naperian
+functor indexed by 8 elements. We define
 %
@@
 combination of \coqe!Rows.T!, \coqe!Cols.T! and \coqe!Slice8.t!
-applied to \coqe!bool! yields a Boolean algebra (over which we can
-therefore write combinational circuits).
+applied to \coqe!bool! yields a Boolean algebra, over which we can
+therefore write combinational circuits.
 
@@
 could also go the other way around, defining an inhabitant of
-\coqe!Cols.T Rows.Ix!. We do not need this construction for \Skinny{}. It
-is necessary in the case of \Gift{}, which proceeds by transposition of
+\coqe!Cols.T Rows.Ix!. We do not need this construction for \Skinny{}
+but it is necessary for \Gift{}, which proceeds by transposition of a
 $4 \times 4$ matrix.
@@
 
-The first step of diffusion process consists in performing a right
-rotation by a specific amount over each individual column. However, we
-do not have enough structure to identify a ``right'' or ``left''
-direction over our containers. The standard tool of the trade to
-achieve this effect is the notion of foldable
-functor~\citep{yorgey:typeclassopedia}:
+The first step of diffusion process consists in applying a right
+rotation over each individual column. However, we do not have enough
+structure to identify a ``right'' or ``left'' direction over our
+containers. The standard tool of the trade to achieve this effect is
+the notion of foldable functor~\citep{yorgey:typeclassopedia}:
 %
@@
 After suitable generalization, this leads us to the following model
-for the operation depicted in Figure~\ref{fig:shiftrows}
+for the \coqe!shiftrows_! operation, depicted in
+Figure~\ref{fig:shiftrows}:
 %
@@
 which represents the first half of the diffusion layer.
-
+%
 %% * Application: mixcolumns
 %% ** figure: psi_ >-> ror 1
-
+%
 The second half corresponds to a combinational circuit \coqe!psi_!
@@
 which exercises the Naperian structure of \coqe!Rows.T! together with
-some underlying Boolean algebra. This is followed by a right rotation
-of rows. Altogether, this defines the operation depicted in Figure~\ref{fig:mixcolumns}
+the underlying Boolean algebra. This is followed by a right rotation
+of rows. Altogether, this defines the \coqe!mixcolumns_! operation
+depicted in Figure~\ref{fig:mixcolumns}
 %

### 2024-10-12 paper.tex
Figure placement
Rephrase

@@
 the underlying Boolean algebra. This is followed by a right rotation
-of rows. Altogether, this defines the \coqe!mixcolumns_! operation
-depicted in Figure~\ref{fig:mixcolumns}
+of rows. Altogether, this defines the \coqe!mixcolumns_! operation,
+depicted in Figure~\ref{fig:mixcolumns}:
 %
@@
 
-\begin{figure}[htp]
+\begin{figure}[tp]
   \centering
@@
 
-At the heart of this layer stands the substitution box \coqe!s8!,
-which is just another combinational circuit (Figure~\ref{fig:s8}):
+At the heart of this layer stands the substitution box \coqe!s8!
+(Figure~\ref{fig:s8}), which is just another combinational circuit:
 %
@@
 These building blocks are chained together to form the confusion
-layer, \coqe!subcells!:
+layer, \coqe!subcells_!:
 %
@@
   \centering
-  \includegraphics[draft=false,height=5cm]{s8}
+  \includegraphics[draft=false,width=.75\textwidth]{s8}
   \caption{Substitution box S8 (from~\citet{beierle:skinny})}
@@
 stored in the smallest addressable memory size, a byte). This would be
-a terrible waste of memory bandwidth and very poor register usage.
+a terrible waste of memory bandwidth and register usage.
 

### 2024-10-12 paper.tex
Rephrase

@@
 would improve memory density. However, we have to be very careful that
-we can still efficiently compute over this packed representation. In
-effect, for \Skinny{} and \Gift{}, this is not quite true: we will have to
-wait until Section~\ref{sec:fixslicing}, which operates on the same
-data representation, to have our cake and eat it too.
+we can still efficiently compute over this packed representation. As
+it turns out, for \Skinny{} and \Gift{}, we will not achieve this just
+now: we will have to wait until Section~\ref{sec:fixslicing}, which
+operates on the same data representation, to have our cake and eat it
+too.
 

### 2024-10-12 paper.tex
Rephrase

--- 2024-10-12 paper.tex
@@
 would improve memory density. However, we have to be very careful that
-we can still efficiently compute over this packed representation. In
-effect, for \Skinny{} and \Gift{}, this is not quite true: we will have to
-wait until Section~\ref{sec:fixslicing}, which operates on the same
-data representation, to have our cake and eat it too.
+we can still efficiently compute over this packed representation. As
+it turns out, for \Skinny{} and \Gift{}, we will not achieve this just
+now: we will have to wait until Section~\ref{sec:fixslicing}, to have
+our cake and eat it too (using the \emph{same} data representation).
 
@@
 blocks at the same time. To this end, we introduce a new Naperian
-functor, \coqe!Double.T!, indexed by the two element set \coqe!Fst!
+functor, \coqe!Double.T!, indexed by the two elements set \coqe!Fst!
 and \coqe!Snd!. This leads to
@@
 \coqfrom{Skinny/BitsliceProof}{invariant_bitslice}
+%
+from their input to their outputs.
 
@@
 off the bitsliced code from the bitsliced types and relational
-invariant. A single round of the cipher thus becomes:
+invariant. A single round of the cipher becomes:
 %
@@
 
-Unfortunately, \coqe!shiftrows_! and \coqe!Rows.ror! involve some very
-fiddly and computationally costly bit twiddling. These amount to
+Unfortunately, \coqe!shiftrows_! and \coqe!Rows.ror! would entail some
+very fiddly and computationally costly bit twiddling. These amount to
 performing bit-level permutations within a machine word. This is a
 frequent pain point when bitslicing the diffusion layer of ciphers: it
-is also true in the case of \Gift{} and \AES{}, for instance. In the next
-section, we present this one weird trick to get permutations to nearly
-vanish.
+is also true in the case of \Gift{} and \AES{}, for instance. In the
+next section, we present this one weird trick to get permutations to
+nearly vanish.
 
@@
 Studying the \Gift{} cipher, \citet{adominicai:fixslicing-gift} noticed that the
-permutation layer can be decompose into, first, a transformation that
-can be efficiently implemented within a register and, second, a
+permutation layer can be decomposed into, first, a transformation that
+can be efficiently implemented over machine words and, second, a
 permutation that commutes with the other layers (confusion and key
@@
   %
-  and can be implemented reasonably efficiently.
+  which can be implemented reasonably efficiently.
 \end{prop}
@@
 Note that we have to transform the input key before hand, so as to
-keep both structures aligned. In particular, this means that we do not
+keep both blocks aligned. In particular, this means that we do not
 have to correct the first round subkey, we will have to apply
@@
 diffusion layer. We observe that putting a state \coqe!phi_ s! into
-the diffusion layer first yield a transformation relativized followed
-by the emission of two \coqe!phi_! steps out of the layer:
+the diffusion layer first yield a transformation relativized to a
+machine word followed by the emission of two \coqe!phi_! steps out of
+the layer:
 %
@@
 absorb two \coqe!phi_! steps. We thus generalize our statement for any
-number of steps, the previous proposition being taken at $n = 1$.
-There are 4 relativized diffusion layers, namely 
+number of \coqe!phi_! steps, the previous statement corresponding to
+the case $n = 1$. There are 4 relativized diffusion layers, namely
 %

### 2024-10-12 paper.tex
Rephrase

@@
 number of \coqe!phi_! steps, the previous statement corresponding to
-the case $n = 1$. There are 4 relativized diffusion layers, namely
+the case $n = 1$:
+%
+\coqfrom{Skinny/Fixslice}{mixcolumns_mod}
+
+Ultimately, there are only 4 relativized diffusion layers, namely
 %
@@
 \coqe!mixcolumns_mod 3!
-%
-\coqfrom{Skinny/Fixslice}{mixcolumns_mod}
 %
@@
 The correctness statement is unsurprising, modulo some light
-bureaucracy to ensure that round subkeys have been suitably generated
+bureaucracy to ensure that round subkeys are in the right format
 %
@@
 We move from one to the next every time we go through a diffusion
-layer, eventually resetting from 3 to 0.
+layer, resetting from the last to the first every 4 steps.
 
@@
 of an efficient implementation. The current form is symbolically
-useful, as it allows us to easily the correctness of fixslicing.
+useful, as it allows us to easily prove the correctness of fixslicing.
 However, it remains a mathematical specification, not a piece of
-software. In effect, we are looking for an open term expressed in the
+software. In effect, we are now looking for a closed expression in the
 language of Boolean algebras (\coqe!xor!, \etc{}), Naperian operations
@@
 
-One solution\footnote{Needless to say, we actually started off with
-the answers, kindly provided by~\citet{adominicai:fixslicing-gift}, and crawled our way
-back to see if an alternative approach would be feasible.}, which
+One solution\footnote{Needless to say, in practice, we actually
+started off from the simplified implementation, kindly provided
+by~\citet{adominicai:fixslicing-gift}, and then used Coq to
+understand how these solutions could have been derived.}, which
 consists in using Coq as a term rewriting engine, is to postulate the
@@
 thanks to the \coqe!set (ident := term)! tactics. We are then left
-with reading off the actual definition of
-\coqe!mixcolumns_mod_explicit!.
+with reading off the definition of \coqe!mixcolumns_mod_explicit! from
+the proof goal.
 

### 2024-10-12 paper.tex
Rephrase

@@
 the origin of the fixslicing
-technique~\citep{adominicai:fixslicing-gift}. \Gift{} is specified in
-terms of the composition
+technique~\citep{adominicai:fixslicing-gift}. \Gift{} works over the type
 %
@@
 records with primitive projections, all the identities boil down to
-the specification, bitsliced forms and fixsliced forms to have the
-same $\beta$-normal $\eta$-long form through the suitable equivalence
+the specification, bitsliced forms and fixsliced forms having the
+same $\beta$-normal $\eta$-long form through the equivalence
 relation!
@@
 kick in: we currently have an implementation defined over \coqe!reg32!
-(and its supporting operations) which we intend to be equivalent to an
+(and its supporting operations) which ought to be equivalent to an
 implementation defined over the type of 32 bits machine words (and its
 supporting operations).
-
+%
 We also wish to pursue the automated generation of fixsliced code,
@@
 diffusion layer.
-
+%
 A litmus test to this project will be our ability to deliver a

### 2024-10-12 paper.tex
typo

@@
 multiplicative inverse over the finite field induced by an
-innocent-looking polynomial. To support such design, the state of
-modern ciphers operate is taken to represent a certain matrix
+innocent-looking polynomial. To support such design, the state modern
+ciphers operate on is taken to represent a certain matrix
 (oftentimes, a 2D or 3D array of bits): the cipher is not specified

### 2024-10-12 paper.tex
typo

@@
 In the long run, our ambition is to offer cryptographers the
-conceptual apparatus to better illuminate their ``Section 5'' scene.
+conceptual apparatus to better illuminate their ``Section 4'' scene.
 Each and every symmetric cryptosystem should not have to exhibit a
@@
 bitsliced designs through sheer intellectual might. It is in fact the
-sole purpose of the traditional ``Section 5'' scene to exhibit a
+sole purpose of the traditional ``Section 4'' scene to exhibit a
 bitsliced witness in all the gory details.
@@
   \end{subfigure}
+  \caption{\Skinny{} diffusion layer}
 \end{figure}
@@
 
-One solution\footnote{Needless to say, in practice, we actually
-started off from the simplified implementation, kindly provided
-by~\citet{adominicai:fixslicing-gift}, and then used Coq to
-understand how these solutions could have been derived.}, which
-consists in using Coq as a term rewriting engine, is to postulate the
-existence of such an explicit form, let us call it
+One solution\footnote{Needless to say, in reality, we first went
+looking find 4 simplified implementations, inspired
+by~\citet{adominicai:fixslicing-gift}. Only later did we use Coq to
+understand how these solutions \emph{could} have been derived from
+first principles.}, which consists in using Coq as a term rewriting
+engine, is to postulate the existence of such an explicit form, let us
+call it
 %

### 2024-10-12 paper.tex
typo

--- 2024-10-12 paper.tex
@@
 If we were to adopt \coqe!state! as our actual representation, the
-memory usage of the resulting program would quite poor. It would take
+memory usage of the resulting program would be quite poor. It would take
 at least 128 bytes to store that many bits (assuming that booleans are
@@
 8 registers in total, imposing a very reasonable amount of register
-pressure to the CPU.
+pressure on the CPU.
 %

### 2024-10-13 paper.tex
typo ?
typo

--- 2024-10-12 paper.tex
@@
 the size of machine words yields nearly twice the throughput, without
-any effort. Evidences tend to suggest that cryptographers obtain
+any effort. Evidence tends to suggest that cryptographers obtain
 bitsliced designs through sheer intellectual might. It is in fact the
@@
 \item we identify the algebraic structure necessary to model the state
-  of modern symmetric cryptography primitives
+  of modern symmetric cryptographic primitives
   (Section~\ref{sec:spec}). Doing so, we delineate a mathematical
@@
 imposes a strict composition order of \coqe!Rows.T! followed by
-\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out the
+\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out of the
 composition. We would thus have the type \coqe!Rows.T (Cols.T bool)!

### 2024-10-13 paper.tex
typo ?
typos

@@
 
-For all intent and purposes, these type constructors shall remain
+For all intents and purposes, these type constructors shall remain
 abstract throughout this article: we interact with them solely through
@@
 non-trivial with the type it is provided as an argument? In other
-word, how to be sure that a \coqe!Rows.T bool! behaves ``similarly''
+words, how to be sure that a \coqe!Rows.T bool! behaves ``similarly''
 as a \coqe!Rows.T nat!? One could try and argue that Coq does not
@@
 More usefully for our purposes, we can derive generic left and right
-rotation rotation operators for any foldable functor \coqe!T! Naperian
+rotation operators for any foldable functor \coqe!T! Naperian
 over a type \coqe!Ix!:
@@
 
-Consequently, our hands are forced when it comes to lifting up
+Consequently, our hands are tied when it comes to lifting up
 individual components:

### 2024-10-14 paper.tex
Yves: missing caption

@@
 multiplicative inverse over the finite field induced by an
-innocent-looking polynomial. To support such design, the state modern
-ciphers operate on is taken to represent a certain matrix
+innocent-looking polynomial. To support such design, the state of
+modern ciphers operate is taken to represent a certain matrix
 (oftentimes, a 2D or 3D array of bits): the cipher is not specified
@@
 the size of machine words yields nearly twice the throughput, without
-any effort. Evidence tends to suggest that cryptographers obtain
+any effort. Evidences tend to suggest that cryptographers obtain
 bitsliced designs through sheer intellectual might. It is in fact the
@@
 \item we identify the algebraic structure necessary to model the state
-  of modern symmetric cryptographic primitives
+  of modern symmetric cryptography primitives
   (Section~\ref{sec:spec}). Doing so, we delineate a mathematical
@@
 
-For all intents and purposes, these type constructors shall remain
+For all intent and purposes, these type constructors shall remain
 abstract throughout this article: we interact with them solely through
@@
 non-trivial with the type it is provided as an argument? In other
-words, how to be sure that a \coqe!Rows.T bool! behaves ``similarly''
+word, how to be sure that a \coqe!Rows.T bool! behaves ``similarly''
 as a \coqe!Rows.T nat!? One could try and argue that Coq does not
@@
 More usefully for our purposes, we can derive generic left and right
-rotation operators for any foldable functor \coqe!T! Naperian
+rotation rotation operators for any foldable functor \coqe!T! Naperian
 over a type \coqe!Ix!:
@@
 
-Consequently, our hands are tied when it comes to lifting up
+Consequently, our hands are forced when it comes to lifting up
 individual components:
@@
 If we were to adopt \coqe!state! as our actual representation, the
-memory usage of the resulting program would be quite poor. It would take
+memory usage of the resulting program would quite poor. It would take
 at least 128 bytes to store that many bits (assuming that booleans are
@@
 imposes a strict composition order of \coqe!Rows.T! followed by
-\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out of the
+\coqe!Cols.T!. Aside from that, we can commute \coqe!Slice8.T! out the
 composition. We would thus have the type \coqe!Rows.T (Cols.T bool)!
@@
 8 registers in total, imposing a very reasonable amount of register
-pressure on the CPU.
+pressure to the CPU.
 %

### 2024-10-14 paper.tex
cryptographic primitive

@@
 As intimidating as cryptographers are, the essence of modern symmetric
-cryptography primitives can in fact be distilled in a few paragraph. A
+cryptographic primitives can in fact be distilled in a few paragraph. A
 primitive can be understood as a purely functional programs. It takes

### 2024-10-14 paper.tex
Dubious humor?

@@
 
-With climate change threatening the Planet, storing only 16 bits in a
-32 bits word would be borderline unethical. We can further increase
-register usage and instantly double throughput by processing two
-blocks at the same time. To this end, we introduce a new Naperian
-functor, \coqe!Double.T!, indexed by the two elements set \coqe!Fst!
-and \coqe!Snd!. This leads to
+% With climate change threatening the Planet, storing only 16 bits in
+% a 32 bits word would be borderline unethical.
+
+Storing only 16 bits in a 32 bits word (on \textsc{Arm} Cortex M)
+remains sub-optimal. We can further increase register usage and
+instantly double throughput by processing two blocks at the same
+time. To this end, we introduce a new Naperian functor,
+\coqe!Double.T!, indexed by the two elements set \coqe!Fst!  and
+\coqe!Snd!. This leads to
 %

### 2024-10-14 paper.tex
Typo

@@
 One solution\footnote{Needless to say, in reality, we first went
-looking find 4 simplified implementations, inspired
-by~\citet{adominicai:fixslicing-gift}. Only later did we use Coq to
-understand how these solutions \emph{could} have been derived from
-first principles.}, which consists in using Coq as a term rewriting
-engine, is to postulate the existence of such an explicit form, let us
-call it
+looking for 4 direct, simplified implementations, inspired
+by~\citet{adominicai:fixslicing-gift}. Then we showed that these
+implementations were equivalent to their respective
+specification. Only later did we use Coq to understand how these
+solutions \emph{could} have been derived from first principles.},
+which consists in using Coq as a term rewriting engine, is to
+postulate the existence of such an explicit form, let us call it
 %

### 2024-10-14 paper.tex
Foldable does not trivially implies rotations

@@
 
-The first step of diffusion process consists in applying a right
+The first step of the diffusion process consists in applying a right
 rotation over each individual column. However, we do not have enough
 structure to identify a ``right'' or ``left'' direction over our
-containers. The standard tool of the trade to achieve this effect is
-the notion of foldable functor~\citep{yorgey:typeclassopedia}:
-%
-\coqfrom{Algebra}{Foldable}
-
-Taking the monoid to be \coqe!list A! (under nil and concatenation,
-\ie{} the free monoid construction), we can read off the
-directionality we were after.
-%
+containers. To do so, we introduce the (regretfully ad-hoc\footnote{We
+were hoping to be able to piggy-back on the notion of
+foldable~\citep{yorgey:typeclassopedia} and Naperian functor to
+identify directions. Intuitively, \coqe!fold! denotes a unique
+left-to-right traversal order. However, we could only turn this into a
+rotation if we asked for a decidable equality over the logarithm of
+the functor. We are not sure yet whether we want to make such a
+commitment.}) notion of circulant functor
+%
+\coqfrom{Algebra}{Circulant}
+%
+taking an \coqe!F!-vector to an \coqe!F! $\times$ \coqe!F! circulant
+matrix (performing a right-rotation of \coqe!F A! at each step) and
+\coqe!F! $\times$ \coqe!F! anticirculant matrix (performing a
+left-rotation of \coqe!F A! at each step).
+
 More usefully for our purposes, we can derive generic left and right
-rotation operators for any foldable functor \coqe!T! Naperian
-over a type \coqe!Ix!:
-%
-\coqfrom{Rows}{ror_sig}
-\coqfrom{Rows}{rol_sig}
-
-\TODO{Foldable to ror/rol}
+rotation operators for any circulant functor \coqe!T!  Naperian over a
+type \coqe!Ix!:
+%
+\coqfrom{Algebra}{ror_rol_sig}
+
+%% The equational theory is unclearas expected for rotations: additive on the
+%% indices when going in the same direction (\coqe!ror i (ror j xs)! and
+%% \coqe!rol i (rol j xs)!), subtractive when going in opposite direction
+%% (\coqe!ror i (rol j xs)! and \coqe!rol i (ror j xs)!).
+
+%\coqfrom{Algebra}{Foldable}
+%% Taking the monoid to be \coqe!list A! (under nil and concatenation,
+%% \ie{} the free monoid construction), we can read off the
+%% directionality we were after.
+%% %
 

### 2024-10-16 paper.tex
Page break
Replace remaining mentions of foldable

@@
   show that modern symmetric ciphers can be modeled through a small
-  set of algebraic structures (Boolean algebra, Naperian and foldable
+  set of algebraic structures (Boolean algebra, Naperian and circulant
   functors). We reveal that these ciphers exhibit some interesting
@@
 language of Boolean algebras (\coqe!xor!, \etc{}), Naperian operations
-(\coqe!init!, \coqe!lookup!) and foldable operations (more precisely,
+(\coqe!init!, \coqe!lookup!) and circulant operations (more precisely,
 \coqe!ror! and \coqe!rol!).
@@
 Making sure that the operations of Boolean algebra, Naperian and
-foldable functors are opaque, $\beta$ reduction yields a term in our
+circulant functors are opaque, $\beta$ reduction yields a term in our
 target language. However, it suffers from significant redundancy: by
@@
 
+\clearpage
+
 \bibliographystyle{unsrtnat}

### 2024-10-16 paper.tex
Typo

@@
 primitive can be understood as a purely functional programs. It takes
-two inputs, the input ``cleartext'' and a list of ``round subkeys''
+two inputs: the ``cleartext'' and a list of ``round subkeys''
 (which are derived from a single key through a non-performance
@@
 each individual round subkeys are processed as ``blocks'' of binary
-data. The block size is of order of a hundred bits (typically, 64, 128
+data. The block size is of the order of a hundred bits (typically, 64, 128
 or 256 bits).

### 2024-10-16 paper.tex
Confusion about diffusion

@@
 
-Following the NSA involvement in the design of the \DES{} diffusion
+Following the NSA involvement in the design of the \DES{} confusion
 layer~\citep{johnson:cold-war-crypto}, cryptographers have been
@@
 harder~\citep{bernstein:something-up-my-sleeves}). For example, the
-diffusion layer of the \AES{} standard is based on the usual
+confusion layer of the \AES{} standard is based on the usual
 multiplicative inverse over the finite field induced by an

### 2024-10-16 paper.tex
Manage national pride
Clarifications

@@
 bitsliced implementation. Bitslicing enables a form of parallel
-execution \emph{within} machine words. We thus gain access to a form
-of CPU-level ``scale out'' to larger machine words, in particular
-through single-instruction multiple-data (SIMD) instructions. Doubling
-the size of machine words yields nearly twice the throughput, without
-any effort. Evidence tends to suggest that cryptographers obtain
-bitsliced designs through sheer intellectual might. It is in fact the
-sole purpose of the traditional ``Section 4'' scene to exhibit a
-bitsliced witness in all the gory details.
+execution \emph{within} machine words, relying sometimes explicitly on
+the availability of vectorized instructions~\citep{beaulieu:speck,
+  bernstein:chacha}. We thus gain access to a form of CPU-level
+``scale out'' to larger machine words, in particular through
+single-instruction multiple-data (SIMD) instructions. Doubling the
+size of machine words yields nearly twice the throughput, without any
+effort. Evidence tends to suggest that cryptographers obtain bitsliced
+designs through sheer intellectual might. It is in fact the sole
+purpose of the traditional ``Section 4'' scene to exhibit a bitsliced
+witness in all the gory details.
 
@@
 We believe that there is also a broader take-away for an audience of
-stubbornly French functional programmers. The present work surfs on
+French functional programmers. The present work surfs on
 the wake of the Squiggol school of programming~\citep{bird:squiggol}.
@@
 peppered with some more structure, so as to 1. unlock algebraic
-manipulation during compilation and 2. exploit parametricity for
+manipulation during compilation and 2. fully exploit parametricity for
 reasoning.
@@
 been a time where we do not have to worry about termination, this is
-it! So we have implemented cryptographic primitive in pure Gallina.
-This is nothing to worry about: after all, more ambitious colleagues
-went from there down to deeply-embedded assembly
-code~\citep{chlipala:coq-crypto}. Besides the ability to prove the
-correctness and, even more usefully, test our code, this also enables
-us to easily prototype a fixslicing compiler, using Coq's
-\texttt{autorewrite} tactics to emulate a transformation-based
-simplification engine~\citep{peyton-jones:haskell-optim} (as part of
-an imaginary optimizing compiler).
+it! So we have implemented cryptographic primitive in pure Gallina,
+following the lead of some of our colleagues who went even further,
+down to deeply-embedded assembly code~\citep{chlipala:coq-crypto}.
+Besides the ability to prove the correctness and, even more usefully,
+test our code, this also enables us to easily prototype a fixslicing
+compiler, using Coq's \texttt{autorewrite} tactics to emulate a
+transformation-based simplification
+engine~\citep{peyton-jones:haskell-optim} (as part of a fictional
+optimizing compiler).
 

### 2024-10-16 paper.tex
Clarify

@@
 \begin{itemize}
-\item we identify the algebraic structure necessary to model the state
-  of modern symmetric cryptographic primitives
+\item we identify the algebraic structure necessary to model modern
+  symmetric cryptographic primitives and, in particular, their state
   (Section~\ref{sec:spec}). Doing so, we delineate a mathematical

### 2024-12-11 paper.tex
Rephrase
Typo
Reviewer B:
  Line 47-48: "a purely functional program~~s~~".
Typo
Reviewer B:
  Line 49: "non-performance-critical".

@@
 cryptographic primitives can in fact be distilled in a few paragraph. A
-primitive can be understood as a purely functional programs. It takes
+primitive can be understood as a purely functional program. It takes
 two inputs: the ``cleartext'' and a list of ``round subkeys''
-(which are derived from a single key through a non-performance
-critical process called ``key schedule'' that we shall ignore here).
+(which are derived from a single key through a non-performance-critical process called ``key schedule'' that we shall ignore here).
 The output is the ``ciphertext''. The cleartext, the ciphertext and
@@
 the design of the whole programming language into this argument.
-
-Instead, we require \coqe!Rows.T! to offer a \coqe!map! operation: it
+%
+Instead, we require \coqe!Rows.T! to offer a \coqe!map! operation: this
 constructively witnesses the fact that a \coqe!Rows.T A! is

### 2024-12-11 paper.tex
Typo
Reviewer B:
  Line 72: "to support such design**s**"

@@
 multiplicative inverse over the finite field induced by an
-innocent-looking polynomial. To support such design, the state modern
+innocent-looking polynomial. To support such designs, the state modern
 ciphers operate on is taken to represent a certain matrix

### 2024-12-11 paper.tex
Add explicit reference
Reviewer B:
  Line 399: "Our previous work on Usuba suggests ..." is there a reference for this work?
Tone down banter
Reviewer B:
  Footnote 3: not useful.

@@
 \coqfrom{Algebra}{Applicative}
+%
+subject to an equational theory expressing the fact that applicative
+functors correspond to lax monoidal
+functors~\citep{maclane:working-categorist} over the Cartesian product
+type. Commutativity of an applicative functor \coqe!F! states that
+arguments can be reduced in any order:
+%
+\coqfrom{Algebra}{ApplicativeComm}
 
@@
 \coqe!Slice8.T!: for any Naperian functor \coqe!F! indexed by
-\coqe!Ix!, there exists a generic permutation operator
+\coqe!Ix!, we can shuffle the elements of a container \coqe!F A! following an arbitrary permutation of indices:
 %
@@
 the next section) as playing, themselves, the role of specifications
-to lower-level refinements. Our previous work on Usuba suggests that
+to lower-level refinements. Our previous work on Usuba~\citep{mercadier:PhD} suggests that
 we will be able to deliver on this promise in the future. Further,
@@
 almost free thanks to the \textsc{Arm} Cortex barrel
-shifter.\footnote{By this remark, we hope to have established our
-credentials as budding cryptographers.}
+shifter.
 
@@
 diffusion layer. We observe that putting a state \coqe!phi_ s! into
-the diffusion layer first yield a transformation relativized to a
+the diffusion layer first yields a transformation relativized to a
 machine word followed by the emission of two \coqe!phi_! steps out of

### 2024-12-11 paper.tex
Define "commutativity" for applicative functors
Reviewer C:
  l. 212 : « commutative », je ne comprends pas ce qui est commutatif ici?
Clarify semantics of `perm`
Reviewer C:
  l. 295 : la permutation est fixée, serait-il possible d'utiliser une permutation quelconque?

--- 2024-12-11 paper.tex
@@
 almost free thanks to the \textsc{Arm} Cortex barrel
-shifter.
+shifter.\footnote{By this remark, we hope to have established our
+credentials as budding cryptographers.}
 

### 2024-12-11 paper.tex
Tone down banter
Reviewer A:
  - p1, l24 “zone à régime restrictive”: comme vous écrivez en anglais,
  il vaut peut-être mieux éviter les blagues qui ne seront pas
  comprises à l'international. C'est parfaitement à propos dans un
  contexte franco-français, mais je ne pense pas que ce soit une
  supposition que vous devriez faire.

-> removed

Reviewer B:

  I find the tone is trying too hard to be unnecessarily goofy / saturday-morning-cartoony
  in places (lines 18-52 for instance), [culminating in Footnote 3 which made me physically cringe.]

-> rewritten

--- 2024-12-11 paper.tex
@@
 
-Anyone who has ever read a cryptography paper knows this for a fact:
-cryptographers are at the top of the academic thought chain. In the
-archetypal symmetric cryptography paper, there is a ``Section 2''
-scene in which cryptographers are bravely waving their most potent
-finite field. In a latter ``Section 4'' scene, the very same
-cryptographers are seen extolling the virtues of some nifty
-\textsc{Arm} Cortex assembly instruction, which takes center stage in
-their hand-crafted, manually optimized implementation. No wonder we
-jealously stash them away in ``zone à régime restrictive''! The rumor
-goes that, behind these padlocked doors, they have developed a set of
-domain-specific techniques to deliver high-throughput software
-implementation of symmetric ciphers. This has been slowly gnawing at
-compiler writers, whose job security is now
-threatened~\citep{bernstein:death-compiler,venkatesh:compiler-security}.
-Luckily for us, every now and then, someone inadvertently let the
-genies out of their Klein bottle and we get to read about the subtle
-art of ``bitslicing''~\citep{biham:DES} and
+Symmetric cryptographic primitives offer a fascinating blend of
+advanced mathematics ---upon which rests their theoretical security
+guarantees--- as well as low-level programming ---upon which rests
+their efficiency but also the absence of side-channel
+attacks~\citep{leroy:side-channel} and resistance to
+fault-injection~\citep{heydemann:fault-attack}.
+%
+In the archetypal symmetric cryptography paper, there is a ``Section
+2'' in which the authors justify the theoretical properties of their
+cipher by appealing to a carefully chosen finite field. In a latter
+``Section 4'', the authors describe their manually optimized
+implementation, usually written directly in
+assembly~\citep{almeida:jasmin}, and report on their performance in
+terms of CPU cycles per encrypted byte.
+%
+Furthermore, cryptographers have developed a set of domain-specific
+techniques to deliver high-throughput software implementation of
+symmetric ciphers, such as ``bitslicing''~\citep{biham:DES} and
 ``fixslicing''~\citep{adominicai:fixslicing-gift,
-  adomnicai:fixslicing-AES-like}.
-
-The present work aims at casting these two notions in the language of
-functional programming. We shall see that bitslicing corresponds to a
-run-off-the-mill \emph{data} representation change (in effect, a
-matrix transposition). But this is only half of the work: our ability
-to systematically transform the \emph{code} operating on such
-representation rests on suitably polymorphic definitions:
-parametricity at its best! Fixslicing, on the other hand, will be
-rationalized through equational reasoning over purely functional
-terms. Identifying commuting expressions is of key importance there:
-programming with algebraic structures (rather than untamed arrays of
-bits) will be instrumental!
+  adomnicai:fixslicing-AES-like}. So far, these techniques have
+remained at odds with off-the-shelf compiler
+technology~\citep{bernstein:death-compiler,venkatesh:compiler-security}.
+
+The present work aims at casting symmetric cryptographic primitives in
+the language of functional programming. In the process, we shall give
+a formal treatment of bitslicing and fixslicing as systematic
+transformations of functional programs. We shall see that bitslicing
+corresponds to a run-off-the-mill \emph{data} representation change
+(in effect, a matrix transposition). But this is only half of the
+work: our ability to systematically transform the \emph{code}
+operating on such representation rests on suitably polymorphic
+definitions: parametricity at its best. Fixslicing, on the other hand,
+will be rationalized through equational reasoning over purely
+functional terms. Identifying commuting expressions is of key
+importance there: programming with algebraic structures (rather than
+untamed arrays of bits) will be instrumental.
 
 In the long run, our ambition is to offer cryptographers the
-conceptual apparatus to better illuminate their ``Section 4'' scene.
+conceptual apparatus to streamline their ``Section 4''.
 Each and every symmetric cryptosystem should not have to exhibit a
@@
 
-As intimidating as cryptographers are, the essence of modern symmetric
-cryptographic primitives can in fact be distilled in a few paragraph. A
+For our present purposes, the essence of modern symmetric
+cryptographic primitives can be distilled in a few paragraph. A
 primitive can be understood as a purely functional program. It takes
-two inputs: the ``cleartext'' and a list of ``round subkeys''
-(which are derived from a single key through a non-performance-critical process called ``key schedule'' that we shall ignore here).
-The output is the ``ciphertext''. The cleartext, the ciphertext and
-each individual round subkeys are processed as ``blocks'' of binary
-data. The block size is of the order of a hundred bits (typically, 64, 128
+two inputs: the ``cleartext'' and a list of ``round subkeys'' (which
+are derived from a single key through a non-performance-critical
+process called ``key schedule'' that we shall ignore here). The output
+is the ``ciphertext''. The cleartext, the ciphertext and each
+individual round subkeys are processed as ``blocks'' of binary data.
+The block size is of the order of a hundred bits (typically, 64, 128
 or 256 bits).

### 2024-12-11 paper.tex
s/cleartext/plaintext/
Christina Boura:
  https://simple.wikipedia.org/wiki/Plaintext
  https://simple.wikipedia.org/wiki/Cleartext

@@
 %% *** thread a `state`
-%% **** initially: clear text
+%% **** initially: plain text
 %% **** finally: cipher text
@@
 primitive can be understood as a purely functional program. It takes
-two inputs: the ``cleartext'' and a list of ``round subkeys'' (which
+two inputs: the ``plaintext'' and a list of ``round subkeys'' (which
 are derived from a single key through a non-performance-critical
 process called ``key schedule'' that we shall ignore here). The output
-is the ``ciphertext''. The cleartext, the ciphertext and each
+is the ``ciphertext''. The plaintext, the ciphertext and each
 individual round subkeys are processed as ``blocks'' of binary data.
@@
 
-Looking at the process of turning the cleartext into a ciphertext, one
+Looking at the process of turning the plaintext into a ciphertext, one
 realizes that cryptographers have been programming in the State monad
 all along: ciphers are described as operating on a ``state'',
-initialized to be the cleartext and read off at the end as the
+initialized to be the plaintext and read off at the end as the
 ciphertext. This state is updated by a ``round'' function that is

### 2024-12-11 paper.tex
Clarify reasonnable expectations
Reviewer A:
  l'introduction me promet qu'elle va éclairer le design des
  algorithmes de cryptographie grâce à l'algèbre, et, finalement, je
  ne me sens que moyennement éclairé de ce point de vu.
  Principalement, l'article me dit qu'en abstrayant au maximum par
  rapport à une algèbre de Boole arbitraire, alors ça simplifie le
  raisonnement sur les algorithmes cryptographiques. Et c'est démontré
  de manière très convaincante. Mais ça ne me donne pas vraiment une
  nouvelle compréhension des algorithmes.

Reviewer C:
  Serait-il possible d'énoncer / prouver

@@
 for fixslicing is next on the list.
+
+Note that the ambitions of our work stops at the level of
+\emph{symmetric} cryptographic \emph{primitives} and, in particular,
+their operational specification. These primitives would typically be
+used (as black-boxes) in cryptographic \emph{protocols}: the analysis
+of such protocols is out of the scope of the present work. Moreover,
+we do not pretend that our conceptual apparatus could be useful to the
+mathematical analysis of these primitives, as performed in a typical
+``Section 2'', since this involves sophisticated, much more high-level
+mathematical objects and results. Our focus is solely on shortening
+the gap between an operational description of a primitive and its
+efficient implementation. In effect, we are designing a programming
+language dedicated to the description of symmetric cryptographic
+primitives.
 

### 2024-12-11 paper.tex
No State monad
Reviewer A:
  p2 l54: “cryptographers have been programming in the State monad” je
  trouve cette affirmation un brin grandilocante quand-même. D'abord,
  il n'y a plus de monade dans le reste de l'article. En plus, si on
  peut effectivement décrire la composition de fonction $A\rightarrow
  A$ comme “programmer dans la monade d'état”, il faut tout de même un
  petit peu de mauvaise fois. En plus, c'est contreproductif pour
  l'argument que la crypto c'est de la programmation fonctionnelle,
  parce qu'il me semble que la composition de fonction, c'est plus ce
  à quoi on pense

@@
 
-Looking at the process of turning the plaintext into a ciphertext, one
-realizes that cryptographers have been programming in the State monad
-all along: ciphers are described as operating on a ``state'',
-initialized to be the plaintext and read off at the end as the
-ciphertext. This state is updated by a ``round'' function that is
-repeated a given number of times (of the order of ten iterations). The
-exact number of iterations is a trade-off between latency (less
-iterations means faster computation) and security (more iterations
-makes for more costly cryptanalysis effort). The round function itself
-is built compositionally from 3 components: a key mixing layer (making
-the encryption process reversible), a confusion layer (smoothing local
+Ciphers are described as operating on a ``state'', initialized to be
+the plaintext and read off at the end as the ciphertext. This state is
+updated by a ``round'' function that is repeated a given number of
+times (of the order of ten iterations). The exact number of iterations
+is a trade-off between latency (less iterations means faster
+computation) and security (more iterations makes for more costly
+cryptanalysis effort). The round function itself is built
+compositionally from 3 components: a key mixing layer (making the
+encryption process reversible), a confusion layer (smoothing local
 correlations among the inputs) and a diffusion layer (spreading

### 2024-12-11 paper.tex
Tone down banter
Justify order of bitsliced state functor
Reviewer A:
  p10, l392: je pense qu'un programmeur typique aura l'intuition que
  la définition de `reg32` va être `Double.T (Rows.T (Cols.T A))`,
  mais ce n'est pas le choix que vous faites. Je réalise que c'est
  l'essence du bitslicing, mais comme vous le remarquez dans votre
  introduction, ce n'est pas une technique particulièrement connue
  hors du milieu de la cryptographie. Je pense qu'il faut expliquer
  pourquoi cette représentation est le bon choix.

@@
 through the fact that \coqe!lookup! and \coqe!init! form a bijection.
-This is also known as a representable functor in categorical
-circles~\citep{maclane:working-categorist}.
+This is also known as a representable functor in the categorical
+literature~\citep{maclane:working-categorist}.
 
@@
 \coqe!Double.T!, indexed by the two elements set \coqe!Fst!  and
-\coqe!Snd!. This leads to
+\coqe!Snd!. Since \coqe!Double.T bool! is a Boolean algebra, we immediately get that \coqe!subcells_! and \coqe!add_shift_rows_! can process two blocks at the same time. This leads to
 %

### 2024-12-11 paper.tex
bugfix: memory consumption of naïve layout
Reviewer A:
  p9, l372: “assuming that booleans are stored in the smallest
  addressable memory size, a byte” vous êtes surprenamment généreux,
  la réalité est généralement bien plus en faveur de votre argument,
  les valeurs sont alignés sur des mots, donc un booléen prend
  typiquement 4 à 8 octets.

En effet, quelle bévue !
About constant-time vs. abstractions
Reviewer A:
  Je m'attendais un aussi à voir une algébraisation, ou du moins une
  abstraction, qui garantie que l'implémentation n'utilise pas
  d'opérateur de contrôle, mais il ne me sembl

@@
 
+Remarkably, reading off the types of \coqe!subcells_!,
+\coqe!add_round_key_!, \coqe!shiftrows_! and \coqe!mixcolumns_!, we
+can argue that this specification supports an implementation immune to
+time-based side-channel attacks. This boils to justifying of the
+absence of control-flow operation depending on cryptographic secrets
+(the plaintext \coqe!s! and, even more crucially, the round keys in
+\coqe!constkey!). It is obvious for \coqe!shiftrows_!, which is
+defined parametrically over any type \coqe!A!. If we can guarantee
+that the Boolean algebra operations can be implemented in
+constant-time (which is the objective of the next Section), then we
+get that the other 3 operations and, therefore, their composition is
+constant-time too.
+
 \section{Bitslicing}
@@
 If we were to adopt \coqe!state! as our actual representation, the
-memory usage of the resulting program would be quite poor. It would take
-at least 128 bytes to store that many bits (assuming that booleans are
-stored in the smallest addressable memory size, a byte). This would be
-a terrible waste of memory bandwidth and register usage.
+memory usage of the resulting program would be quite poor. It would
+take at least 512 bytes to store these 128 Booleans (assuming that each
+boolean is stored in the smallest addressable memory size, a 4-bytes
+word). This would be a terrible waste of memory bandwidth and register
+usage.
 
@@
 grouping individual booleans into a single machine word. Doing so
-would improve memory density. However, we have to be very careful that
-we can still efficiently compute over this packed representation. As
-it turns out, for \Skinny{} and \Gift{}, we will not achieve this just
-now: we will have to wait until Section~\ref{sec:fixslicing}, to have
-our cake and eat it too (using the \emph{same} data representation).
+would improve memory density. This would also lead to an
+implementation of the Boolean algebra that operates uniformly over
+whole machine words, using machine instructions to process an entire
+slice in a set number of CPU cycles. This is the fundamental
+assumption behind bitslicing~\citep{biham:DES}: applying, say, a
+logical \coqe!AND! machine instruction over two registers (each
+representing a slice of the cipher state) takes one CPU cycle and
+therefore takes the same amount of time, independently of the values
+stored in the registers\footnote{A careful reader will have noted the
+non sequitur: taking one CPU cycle does not actually imply
+taking identitical (physical) execution time~\citep{wang:hertzbleed}.
+In very specific circumstances where the cryptographic primitive acts as a signal amplifier, this can actually be exploited: we do not account for this sort of second-order effect in the present work.}.
+
+%% However, we have to be very careful that
+%% we can still efficiently compute over this packed representation. As
+%% it turns out, for \Skinny{} and \Gift{}, we will not achieve this just
+%% now: we will have to wait until Section~\ref{sec:fixslicing}, to have
+%% our cake and eat it too (using the \emph{same} data representation).
 

### 2024-12-11 paper.tex
About equality
Reviewer A:
  Quelque chose que j'ai trouvé un peu trop implicite, en terme de
  description de la formalisation, c'est sur le traitement des
  égalités. Je devine que les algèbres de Bool que vous considérez
  doivent être des algèbres de Bool par rapport à l'égalité native de
  Coq, pas de quotient. C'est un peu limitant, mais ça n'a pas l'air
  de vous poser de problème en pratique. J'aurais probablement aimé un
  peu plus de commentaire là-dessus.

@@
 functors~\citep{maclane:working-categorist} over the Cartesian product
-type. Commutativity of an applicative functor \coqe!F! states that
-arguments can be reduced in any order:
+type. Commutativity of an applicative functor \coqe!F!
+states\footnote{For conceptual simplicity, we shall state our
+properties using an extensional equality throughout this article: two
+functions can be substituted for one another ---they are equal--- if
+they produce the same outputs given the same inputs. Given our focus
+on terminating functions (cryptographic primitives) from finite types
+to finite types (blocks of plaintext to blocks ciphertext), this is
+perfectly justified, even in an intuitionist type-theory such as the
+Calculus of Inductive Constructions of Coq.} that arguments can be
+reduced in any order:
 %

### 2024-12-11 paper.tex
Clarify meaning of phi^4 ~ id
Reviewer C:
  dans la partie fixslicing, je ne comprends pas l'utilité d'enchaîner
  les itérations plusieurs dizaines de fois, si 4 itérations
  correspondent plus où moins à l'identité.

@@
 
+In effect, we have transformed the bitsliced round from
+Section~\ref{sec:bitslicing}, which applied \coqe!phi_! in every round
+over a uniform state representation, into 4 specialized rounds that
+each operate over a specific layout of the state. We have thus traded
+code size (specializing the round function to 4 specific layouts) for
+improved run-time performance (avoiding the bit-twiddling operations
+implied by \coqe!phi_!).
+
 %% * Round

### 2024-12-11 paper.tex
Circulant functor: alternative characterization?
Rephrase technical motivation for fixslicing
Reviewer A:

  p12, l494: “Being able to postpots `phi_`”: c'est la première fois,
  me semble-t-il, que vous parlez de retarder l'application de `phi_`,
  mais vous en parlez comme si c'était déjà un savoir partagé avec le
  lecteur, alors que vous n'introduisez l'idée que dans la proposition
  3 un peu plus bas.

-> en effet, bug dans l'écriture / les relectures

@@
 
+%% JFLA'25, reviewer A:
+%% - p6, l240: de l'abstraction `Circulant`. En fait, ce que fait
+%%   Circulant pour vous, c'est exactement définir un isomorphisme entre
+%%   vos ensembles d'indexe et un groupe cyclique
+%%   $\mathbb{Z}/n\mathbb{Z}$, je pense que vous vous éviteriez pas mal
+%%   de complication en fait si vous vous autorisiez à penser en ces
+%%   termes. Ensuite, vous pouvez faire tout ce que vous voulez en termes
+%%   d'`ini`, `index`, et cette structure de groupe. Même les _applicative_
+%%   (qui sont tout de même plus naturels) ça ne vous sert
+%%   qu'essentiellement une fois. (sinon, il me semble que votre
+%%   `Circulant` identifie deux structures de comonades qui sont
+%%   compatibles avec la même counité (`index _ 0`), mais je ne sais pas
+%%   si c'est important).
+
 More usefully for our purposes, we can derive generic left and right
@@
 
-Being able to postpone \coqe!phi_! would do us no good if we were
-nonetheless forced to run it anyway. The second, crucial observation
-is that, after only 4 steps, \coqe!phi_! is almost an identity.
-
+Fixslicing rests on the observation that after 4 steps, \coqe!phi_! is
+almost an identity:
+%
 \begin{prop}
@@
 In typical mode of operation, \Skinny{} will perform between 32 and 56
-rounds. Being able to accumulate \coqe!phi_! across chunks of 4
-iterations, we can effectively make them disappear, leaving only a
-\coqe!tau_! step every four iterations.
+rounds. If we can postpone the application of \coqe!phi_! across 4
+rounds, we could effectively make them disappear, leaving only a
+\coqe!tau_! step every four rounds.
 

### 2024-12-11 paper.tex
Be (somewhat) more explicit about parametricity
Reviewer A:

  Vous parlez de paramétricité à plusieurs endroit. Mais je n'ai pas
  vraiment compris ce que vous faites avec? Est-ce que vous prouvez
  des lemmes de paramétricités pour vos fonctions? Peut-être en
  utilisant un plugin (je sais que des plugins pour prouver la
  paramétricité de fonctions polymorphes automatiquement, mais je n'en
  sais pas l'état). C'est un point qui mériterait d'être éclairci,
  probablement.

@@
 The relational invariant will then naturally and compositionally
-percolate through the various layers of the ciphers. We can then read
+percolate through the various layers of the ciphers, following the parametricity interpretation of types~\citep{lasson:param}. We can then read
 off the bitsliced code from the bitsliced types and relational

### 2024-12-12 paper.tex
plural

@@
 
-Note that the ambitions of our work stops at the level of
+Note that the ambitions of our work stop at the level of
 \emph{symmetric} cryptographic \emph{primitives} and, in particular,
@@
 For our present purposes, the essence of modern symmetric
-cryptographic primitives can be distilled in a few paragraph. A
+cryptographic primitives can be distilled in a few paragraphs. A
 primitive can be understood as a purely functional program. It takes

### 2024-12-12 paper.tex
less -> fewer ?

@@
 times (of the order of ten iterations). The exact number of iterations
-is a trade-off between latency (less iterations means faster
+is a trade-off between latency (fewer iterations means faster
 computation) and security (more iterations makes for more costly

### 2024-12-12 paper.tex
plural

@@
 
-To avoid side-channel attack (for example, based on
+To avoid side-channel attacks (for example, based on
 timing~\citep{bernstein:timing-attack-AES}) over software

### 2024-12-12 paper.tex
suboptimal
before hand -> beforehand

@@
 Storing only 16 bits in a 32 bits word (on \textsc{Arm} Cortex M)
-remains sub-optimal. We can further increase register usage and
+remains suboptimal. We can further increase register usage and
 instantly double throughput by processing two blocks at the same
@@
 
-Note that we have to transform the input key before hand, so as to
+Note that we have to transform the input key beforehand, so as to
 keep both blocks aligned. In particular, this means that we do not

### 2024-12-17 paper.tex
Tone down banter

@@
 the wake of the Squiggol school of programming~\citep{bird:squiggol}.
-Because our government-approved programming language does not yet
+Because OCaml, our favorite  programming language, does not yet
 support ad-hoc polymorphism~\citep{white:modular-implicit}, the French
@@
 is also true in the case of \Gift{} and \AES{}, for instance. In the
-next section, we present this one weird trick to get permutations to
+next section, we present a clever trick to get permutations to
 nearly vanish.
